---
title: "Using Weights in GLLAMMR"
author: "GLLAMMR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Weights in GLLAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

Note: the code in this vignette is illustrative and is not evaluated when the vignette is built.

GLLAMMR supports two types of weights for all model families:

- **Frequency weights (fweights)**: Integer weights representing the number of observations at each unique covariate pattern
- **Probability weights (pweights)**: Continuous weights for survey sampling or inverse probability weighting

This vignette demonstrates how to use weights with various model types in GLLAMMR.

## Types of Weights

### Frequency Weights

Frequency weights are used when your data represents aggregated observations. Each row represents multiple observations with identical covariate values.

**Example:** Instead of 100 individual observations, you have 10 rows where each row represents 10 identical observations.

```{r fweights_example, eval=FALSE}
# Original data (100 observations)
# x: 1,1,1,1,1,2,2,2,2,2,...

# Aggregated data (10 observations)
data_agg <- data.frame(
  x = rep(1:10, each = 1),
  y = c(5, 8, 6, 7, 9, 4, 5, 8, 6, 7),
  freq = rep(10, 10)  # Each row represents 10 observations
)
```

### Probability Weights

Probability weights are used in survey sampling where different observations have different sampling probabilities, or in causal inference with inverse probability weights.

**Example:** Survey data where some demographic groups are oversampled.

```{r pweights_example, eval=FALSE}
# Survey data with sampling weights
data_survey <- data.frame(
  outcome = rbinom(1000, 1, 0.5),
  treatment = rbinom(1000, 1, 0.5),
  group = factor(sample(1:10, 1000, replace = TRUE)),
  weight = runif(1000, 0.5, 2.0)  # Inverse of sampling probability
)
```

## Using Weights in GLMM

All GLMM families (Gaussian, Binomial, Poisson) support weights.

### Example 1: Binomial GLMM with Aggregated Data

```{r binomial_weights, eval=FALSE}
library(GLLAMMR)

# Aggregated binary data
data <- data.frame(
  success = c(15, 23, 18, 20, 25),
  total = c(20, 30, 25, 25, 30),
  treatment = c(0, 0, 1, 1, 1),
  clinic = factor(1:5)
)

# Create frequency weights
data$freq <- data$total

# Expand to binary format
data$prop <- data$success / data$total

# Fit with weights
fit <- gllamm(cbind(success, total - success) ~ treatment + (1 | clinic),
              data = data,
              family = binomial(),
              weights = data$freq)

summary(fit)
```

### Example 2: Poisson GLMM with Survey Weights

```{r poisson_weights, eval=FALSE}
# Survey data with counts
data <- data.frame(
  count = rpois(100, lambda = 5),
  x1 = rnorm(100),
  group = factor(rep(1:10, each = 10)),
  survey_weight = runif(100, 0.8, 1.5)
)

# Fit Poisson model with weights
fit <- gllamm(count ~ x1 + (1 | group),
              data = data,
              family = poisson(),
              weights = data$survey_weight)

summary(fit)
```

## Using Weights in IRT

IRT models support weights at the person level. Weights are automatically expanded to the item-response level.

### Example: IRT with Weighted Persons

```{r irt_weights, eval=FALSE}
# Item response matrix
n_persons <- 200
n_items <- 10
responses <- matrix(rbinom(n_persons * n_items, 1, 0.6),
                   n_persons, n_items)

# Person weights (e.g., sampling weights)
person_weights <- runif(n_persons, 0.5, 2.0)

# Fit weighted 2PL model
fit_irt <- fit_irt(responses,
                   model = "2PL",
                   weights = person_weights)

# Item parameters account for differential person weighting
print(fit_irt$item_parameters)
```

**Key point:** The weights vector has length `n_persons`, but internally GLLAMMR expands it to length `n_persons × n_items` for the TMB likelihood.

### Example: EIRT with Weights

```{r eirt_weights, eval=FALSE}
# Item data
item_data <- data.frame(
  item_id = 1:n_items,
  difficulty_pred = rnorm(n_items)
)

# Fit weighted EIRT model
fit_eirt <- fit_eirt(responses,
                     item_data = item_data,
                     difficulty_formula = ~ difficulty_pred,
                     model = "2PL",
                     weights = person_weights)

summary(fit_eirt)
```

## Using Weights in Ordinal Models

Ordinal models support weights for observations with ordinal outcomes.

```{r ordinal_weights, eval=FALSE}
# Ordinal data
data <- data.frame(
  rating = sample(1:5, 500, replace = TRUE),
  temp = rnorm(500),
  judge = factor(rep(1:25, each = 20)),
  weight = runif(500, 0.8, 1.2)
)

# Fit proportional odds model with weights
fit_ord <- gllamm(rating ~ temp + (1 | judge),
                  data = data,
                  family = ordinal(link = "logit"),
                  weights = data$weight)

summary(fit_ord)
```

## Using Weights in Latent Class Analysis

LCA supports weights for case-level observations.

```{r lca_weights, eval=FALSE}
# Binary indicators for LCA
indicators <- matrix(rbinom(500 * 8, 1, 0.6), 500, 8)

# Case weights
case_weights <- runif(500, 0.5, 1.5)

# Fit weighted LCA
fit_lca <- fit_lca(indicators,
                   nclass = 3,
                   weights = case_weights)

# Class proportions account for weighting
print(fit_lca$class_proportions)
```

## Using Weights in Multinomial Models

```{r multinomial_weights, eval=FALSE}
# Multinomial choice data
data <- data.frame(
  choice = sample(0:2, 300, replace = TRUE),
  price = rnorm(300),
  quality = rnorm(300),
  person = factor(rep(1:100, each = 3)),
  weight = runif(300, 0.7, 1.3)
)

# Fit weighted multinomial model
fit_mult <- fit_multinomial(choice ~ price + quality + (1 | person),
                            data = data,
                            weights = data$weight)

summary(fit_mult)
```

## Using Weights in Survival Models

Survival models support weights for time-to-event data.

```{r survival_weights, eval=FALSE}
# Survival data
data <- data.frame(
  time = rexp(200, rate = 0.1),
  status = rbinom(200, 1, 0.7),
  treatment = rep(0:1, each = 100),
  hospital = factor(rep(1:10, each = 20)),
  weight = runif(200, 0.6, 1.4)
)

# Fit weighted survival model
fit_surv <- fit_survival(time ~ treatment + (1 | hospital),
                        data = data,
                        event = data$status,
                        distribution = "Weibull",
                        weights = data$weight)

summary(fit_surv)
```

## Weight Validation

GLLAMMR automatically validates weights:

```{r validation, eval=FALSE}
# These will produce errors:

# Negative weights
fit <- gllamm(y ~ x + (1 | g), data = data,
              weights = c(-1, 2, 3))  # Error!

# Wrong length
fit <- gllamm(y ~ x + (1 | g), data = data,
              weights = c(1, 2))  # Error if nrow(data) != 2

# Missing values
fit <- gllamm(y ~ x + (1 | g), data = data,
              weights = c(1, NA, 3))  # Error!
```

All weights must be:
- Non-negative
- Same length as number of observations
- Non-missing (no NA values)

## Practical Guidelines

### When to Use Frequency Weights

1. **Aggregated data**: You have summary statistics rather than individual observations
2. **Computational efficiency**: Reduce dataset size when many observations are identical
3. **Count data**: Each row represents a count of occurrences

### When to Use Probability Weights

1. **Survey sampling**: Observations sampled with unequal probabilities
2. **Inverse probability weighting**: Causal inference adjusting for treatment assignment
3. **Meta-analysis**: Combining studies with different sample sizes
4. **Stratified sampling**: Correcting for oversampling of subgroups

### Best Practices

1. **Document your weights**: Clearly explain what the weights represent
2. **Check diagnostics**: Weighted models may have different convergence properties
3. **Report weighted sample size**: For frequency weights, report effective N
4. **Validate results**: Compare weighted vs unweighted estimates when possible
5. **Center weights**: Consider centering probability weights around 1

## Implementation Details

### How Weights Affect the Likelihood

In the TMB template, weights multiply the log-likelihood contribution:

```cpp
// Without weights
nll -= log(prob);

// With weights
Type w_i = weights(i);
nll -= w_i * log(prob);
```

This means:
- Observations with weight = 2 contribute twice as much to the likelihood
- Observations with weight = 0.5 contribute half as much
- Observations with weight = 0 are effectively excluded

### Computational Cost

Weights add minimal computational overhead:
- Validation: O(n) where n is number of observations
- Likelihood evaluation: Single multiplication per observation
- No change to optimization algorithm

## Comparison with Other Software

| Software | Frequency Weights | Probability Weights |
|----------|-------------------|---------------------|
| GLLAMMR  | ✓ | ✓ |
| lme4     | ✓ (via cbind) | Limited |
| glmmTMB  | ✓ | ✓ |
| Stata    | ✓ (fweight) | ✓ (pweight) |
| SAS      | ✓ (FREQ) | ✓ (WEIGHT) |

GLLAMMR provides unified support across all model families.

## Summary

- All GLLAMMR model families support weights
- Use `weights` parameter in all fitting functions
- Weights are validated automatically
- Choose frequency vs probability weights based on data structure
- For IRT models, provide person-level weights (auto-expanded internally)

For more examples and technical details, see:
- `?gllamm` for GLMM weights
- `?fit_irt` for IRT weights
- `?fit_ordinal` for ordinal model weights

## References

- Pfeffermann, D. (1993). The role of sampling weights when modeling survey data. *International Statistical Review*, 61(2), 317-337.

- DuMouchel, W. H., & Duncan, G. J. (1983). Using sample survey weights in multiple regression analyses of stratified samples. *Journal of the American Statistical Association*, 78(383), 535-543.
