---
title: "Cognitive Diagnosis Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cognitive Diagnosis Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## From a continuous trait to attribute profiles

IRT summarizes a person with one number. Cognitive diagnosis models
(CDMs) replace that with a *profile* of binary skills: can this student
subtract fractions, find common denominators, simplify? The latent
classes are the $2^A$ attribute profiles, and a **Q-matrix** declares
which skills each item requires. The payoff is diagnostic: instead of a
score, each person gets a posterior probability of mastering each skill.

`fit_cdm()` (equivalently `gllamm(Y, family = cdm(Q, ...))`) implements
the saturated **G-DINA** family with **DINA** and **DINO** as special
cases, estimated by a compiled, accelerated EM and validated against the
CDM package (de la Torre's fraction-subtraction data: identical item
parameters; simulated G-DINA: log-likelihoods equal to 1e-4).

## Simulating a diagnostic assessment

Three skills, 15 items: single-skill items plus items requiring two or
all three skills (conjunctively — the DINA rule).

```{r simulate}
library(gllammr)
set.seed(2026)
A <- 3
Q <- rbind(diag(3), diag(3), diag(3),
           c(1, 1, 0), c(0, 1, 1), c(1, 0, 1),
           c(1, 1, 0), c(0, 1, 1), c(1, 1, 1))
colnames(Q) <- c("subtract", "denominator", "simplify")
n <- 1000
alpha <- matrix(rbinom(n * A, 1, 0.55), n, A)   # true skill profiles
guess <- runif(nrow(Q), 0.05, 0.20)
slip  <- runif(nrow(Q), 0.05, 0.20)
eta <- sapply(seq_len(nrow(Q)), function(j) {
  as.integer(rowSums(alpha[, Q[j, ] == 1, drop = FALSE]) == sum(Q[j, ]))
})
Y <- sapply(seq_len(nrow(Q)), function(j) {
  rbinom(n, 1, ifelse(eta[, j] == 1, 1 - slip[j], guess[j]))
})
```

## Fitting the DINA model

DINA is conjunctive: you answer correctly (beyond guessing) only if you
master *every* required skill. Each item gets a guessing and a slip
parameter.

```{r dina}
fit <- fit_cdm(Y, Q, model = "dina")
print(fit)
```

Parameter recovery is direct to check here:

```{r recovery}
ghat <- sapply(fit$item_params, function(e) e$guess)
shat <- sapply(fit$item_params, function(e) e$slip)
round(rbind(guess_error = max(abs(ghat - guess)),
            slip_error = max(abs(shat - slip))), 3)
mean(fit$modal_attributes == alpha)   # per-skill classification accuracy
```

## The diagnostic output

The clinically useful quantity is the person-by-skill mastery posterior:

```{r posteriors}
head(round(fit$attribute_posteriors, 3))
```

Each row is one student; each entry is P(mastered that skill | their
responses). Profile prevalences (`fit$profile_probs`) describe the
population.

## G-DINA: when conjunction is too strict

The saturated G-DINA (the default, `model = "gdina"`) frees one response
probability per *combination* of required skills — partial mastery can
partially help. Because DINA is nested in G-DINA, a likelihood-ratio
comparison asks whether the conjunctive restriction holds:

```{r gdina}
fit_g <- fit_cdm(Y, Q, model = "gdina")
c(dina = fit$logLik, gdina = fit_g$logLik,
  lr = 2 * (fit_g$logLik - fit$logLik))
```

A small statistic relative to the extra parameters (here the data are
truly DINA) supports the simpler model. `model = "dino"` gives the
disjunctive counterpart (any one skill suffices).

Item kernels under G-DINA read directly as P(correct) per reduced skill
pattern — for a two-skill item, four probabilities:

```{r kernel}
fit_g$item_params[[10]]
```

## Monotonicity

By default (`monotone = TRUE`) mastering more of an item's skills can
never *lower* the success probability — enforced by isotonic regression
over the skill lattice in the M-step, the same machinery behind
gllammr's order-restricted latent class models. Set `monotone = FALSE`
for the unconstrained saturated model; the constrained fit can never
beat it on likelihood, and a large gap flags items whose kernels violate
the ordering.

## Attribute hierarchies

If skills have prerequisites — you cannot simplify fractions before you
can subtract them — the profile space shrinks. Declare prerequisite
pairs and impossible profiles are removed:

```{r hierarchy}
fit_h <- fit_cdm(Y, Q, model = "dina",
                 hierarchy = list(c("subtract", "simplify")))
fit_h$profile_labels    # profiles with simplify-but-not-subtract are gone
```

Linear, convergent, and divergent hierarchies are all expressible as
pair lists; cycles are rejected.

## Connection to the GLLAMM framework

A CDM is a structured latent class model: with one attribute, the
unconstrained G-DINA *is* a two-class LCA (gllammr's test suite verifies
the log-likelihoods agree to 1e-11). The Q-matrix supplies the equality
structure (items depend only on their measured skills) and monotonicity
the order structure — both restrictions of the GLLAMM parameter space,
not departures from it. See `vignette("latent-class")` for the
unstructured and order-restricted latent class models this builds on.

## References

- de la Torre, J. (2011). The generalized DINA model framework.
  *Psychometrika*, 76, 179–199.
- Junker, B. W., & Sijtsma, K. (2001). Cognitive assessment models with
  few assumptions. *Applied Psychological Measurement*, 25, 258–272.
- Rupp, A. A., Templin, J., & Henson, R. A. (2010). *Diagnostic
  Measurement: Theory, Methods, and Applications*. Guilford.
