---
title: "Explanatory IRT: the De Boeck & Wilson Framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explanatory IRT: the De Boeck & Wilson Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = requireNamespace("lme4", quietly = TRUE)
)
```

## Items as outcomes, not just instruments

Classical IRT treats each item as a separate entity with its own freely
estimated parameters. De Boeck & Wilson (2004, *Explanatory Item
Response Models*) reframed the field: an IRT model is a generalized
linear mixed model, and once you see it that way, item properties become
*predictors*. Instead of asking "how difficult is item 7?", you ask
"*why* is item 7 difficult — and how well do its design features explain
that?"

gllammr implements this framework in `fit_eirt()` (equivalently
`gllamm(resp, family = eirt(...))`). This vignette reproduces the
canonical analysis: the **verbal aggression data** (316 persons x 24
items, shipped with lme4), where each item is built from a 2 x 2 x 3
design — behaviour mode (want/do), situation type (other/self-to-blame),
and behaviour type (curse/scold/shout).

```{r data}
library(gllammr)
data("VerbAgg", package = "lme4")

# Persons x items response matrix (1 = "yes")
VerbAgg$y <- as.integer(VerbAgg$r2 == "Y")
resp <- with(VerbAgg, tapply(y, list(id, item), identity))
resp <- matrix(as.integer(resp), nrow = nrow(resp),
               dimnames = dimnames(resp))

# Item design factors, one row per item
item_info <- unique(VerbAgg[, c("item", "btype", "situ", "mode")])
item_info <- item_info[match(colnames(resp),
                             as.character(item_info$item)), ]
item_data <- data.frame(btype = factor(item_info$btype, ordered = FALSE),
                        situ  = factor(item_info$situ, ordered = FALSE),
                        mode  = factor(item_info$mode, ordered = FALSE))
str(item_data)
```

## The LLTM: difficulty explained by design

The linear logistic test model (LLTM) replaces the 24 free item
difficulties with a regression on the three design factors:
$b_i = \mathbf{w}_i'\boldsymbol\gamma$. Five parameters instead of 24.

```{r lltm}
lltm <- fit_eirt(resp, item_data,
                 difficulty_formula = ~ btype + situ + mode,
                 model = "Rasch", item_residuals = FALSE)
lltm$regression_coefficients$difficulty
```

Positive coefficients mean *harder to endorse*: shouting is about two
logits harder to admit to than cursing, self-blame situations are about
one logit harder than other-blame, and actually *doing* is harder than
merely *wanting to*.

## LLTM + error: random item residuals

The pure LLTM assumes the design explains item difficulty *perfectly* —
almost never true. The LLTM-plus-error model adds a random item residual,
$b_i = \mathbf{w}_i'\boldsymbol\gamma + \epsilon_i$, which is the
`fit_eirt()` default:

```{r lltm-error}
lltm_e <- fit_eirt(resp, item_data,
                   difficulty_formula = ~ btype + situ + mode,
                   model = "Rasch", item_residuals = TRUE)
lltm_e$regression_coefficients$difficulty
lltm_e$residual_sd$difficulty
```

The residual SD (~0.34) against design effects of 0.7–2.1 logits says
the three facets explain *most* — but not all — of the difficulty
variation. The likelihood-ratio statistic between the two models makes
the point formally:

```{r lr}
2 * (lltm_e$logLik - lltm$logLik)
```

## The GLMM connection, verified

The LLTM-plus-error is *exactly* a binomial GLMM with crossed person and
item random effects — De Boeck & Wilson's central insight. You can check
this inside gllammr itself: an intercept-only explanatory model equals
the crossed-effects GLMM through the unified interface (and both equal
`lme4::glmer`).

```{r crosswalk}
e0 <- fit_eirt(resp, data.frame(int = rep(1, 24)),
               difficulty_formula = ~ 1, model = "Rasch",
               item_residuals = TRUE)
g0 <- gllamm(y ~ 1 + (1 | id) + (1 | item), data = VerbAgg,
             family = binomial())
c(eirt = e0$logLik, gllamm = g0$logLik)
```

Identical marginal likelihoods: the explanatory IRT model *is* a GLLAMM.

## Polytomous items: the Kim & Wilson extension

The original responses have three categories (no / perhaps / yes).
Kim & Wilson (2019) extended item-explanatory models to polytomous data;
their "location-explanatory" partial credit model is `model = "PCM"`
with the same difficulty formula. (Their reference categories are
want / self / shout, so we relevel.)

```{r pcm}
VerbAgg$y3 <- as.integer(VerbAgg$resp)   # 1 = no, 2 = perhaps, 3 = yes
resp3 <- with(VerbAgg, tapply(y3, list(id, item), identity))
resp3 <- matrix(as.integer(resp3), nrow = nrow(resp3),
                dimnames = dimnames(resp3))
item_kw <- data.frame(
  btype = relevel(item_data$btype, ref = "shout"),
  situ  = relevel(item_data$situ, ref = "self"),
  mode  = item_data$mode)

pcm_e <- fit_eirt(resp3, item_kw,
                  difficulty_formula = ~ mode + situ + btype,
                  model = "PCM", item_residuals = TRUE)
pcm_e$regression_coefficients$difficulty
```

These reproduce Kim & Wilson's published "MFRM + OIE" estimates
(intercept 1.69, Do 0.49, Other −0.89, Curse −1.38, Scold −0.70) within
a few hundredths — their numbers are Bayesian posterior means, ours
maximum likelihood. Step-specific item properties (their LPCM) are
available through `threshold_formula`.

## Covariates at the step level

Two distinct kinds of "step predictor" arise in practice, and the
package separates them:

- `threshold_formula`: **item properties whose effects differ by step**.
  Each covariate gets one coefficient per threshold, constrained to sum
  to zero across an item's thresholds — they are step-specific
  *deviations* around the item-level effect, which belongs in
  `difficulty_formula`. (The two are jointly identified this way.)
- `step_formula` with `step_data`: **covariates that vary within an item
  across its steps**, each with a single common coefficient — e.g. each
  scoring step of a rubric demands a different skill. `step_data` has
  one row per item-step cell in item-major order.

The full decomposition is
$\delta_{im} = b_i + \xi_{0m} + \sum_k \eta_k x_{imk} + e_{im}$:
item location from the difficulty regression, per-step baselines,
step-level covariate effects, and (with `item_residuals = TRUE`)
step residuals.

```{r step-level}
# Suppose each item-step is rated for the cognitive demand of that step
set.seed(1)
step_props <- data.frame(
  demand = rnorm(ncol(resp3) * 2)   # 3 categories -> 2 steps per item
)
pcm_step <- fit_eirt(resp3, item_kw,
                     difficulty_formula = ~ mode + situ + btype,
                     step_formula = ~ demand, step_data = step_props,
                     model = "PCM", item_residuals = TRUE)
pcm_step$regression_coefficients$step
```

A step covariate that happens to be constant within items is exactly
equivalent to entering it in `difficulty_formula`; genuinely redundant
cross-level specifications are rejected with a rank-deficiency error
rather than fit on a ridge.

## Explaining discrimination too

For models with a discrimination parameter (2PL, GRM, GPCM),
`discrimination_formula` regresses the *log* discrimination on item
properties: $a_i = \exp(\mathbf{w}_i'\boldsymbol\delta + \epsilon_{a,i})$,
so coefficients are multiplicative effects.

```{r discrimination}
gpcm_e <- fit_eirt(resp3, item_kw,
                   difficulty_formula = ~ mode + situ + btype,
                   discrimination_formula = ~ mode,
                   model = "GPCM", item_residuals = TRUE)
exp(gpcm_e$regression_coefficients$discrimination)
```

"Do" items discriminate about 30% better than "want" items: behavioural
commitment separates people more sharply than hypothetical inclination.

## Relation to the GLLAMM framework

Difficulties enter the linear predictor linearly — canonical GLLAMM.
Discriminations are GLLAMM factor loadings modelled on the log scale, a
standard positivity-preserving reparameterization. Multilevel person
structures (`random = ~ (1 | school)`) and person-level random slopes
compose with all of the above; see `vignette("multilevel-irt")`.

## References

- De Boeck, P., & Wilson, M. (2004). *Explanatory Item Response Models:
  A Generalized Linear and Nonlinear Approach*. Springer.
- Kim, J., & Wilson, M. (2019). Polytomous item explanatory IRT models
  with random item effects. *Measurement*, 151, 107062.
- Fischer, G. H. (1973). The linear logistic test model as an instrument
  in educational research. *Acta Psychologica*, 37, 359–374.
