---
title: "Getting Started with GLLAMMR"
author: "GLLAMMR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with GLLAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE when examples can run
)
```

## Introduction

GLLAMMR (Generalized Linear Latent and Mixed Models in R) provides a comprehensive framework for fitting multilevel latent variable models. This vignette introduces the basic functionality.

## Installation

```{r install}
# Install from GitHub
# remotes::install_github("yourusername/GLLAMMR")

library(GLLAMMR)
```

## Basic Random Intercept Model

The simplest GLLAMM model is a random intercept model:

$$y_{ij} = \beta_0 + \beta_1 x_{ij} + u_j + \epsilon_{ij}$$

where:
- $y_{ij}$ is the response for observation $i$ in group $j$
- $x_{ij}$ is a predictor
- $u_j \sim N(0, \sigma_u^2)$ is the random intercept for group $j$
- $\epsilon_{ij} \sim N(0, \sigma^2)$ is the residual error

### Simulate Data

```{r simulate-basic}
set.seed(123)

# Parameters
n_groups <- 30
n_per_group <- 10
beta_0 <- 2.0
beta_1 <- -0.5
sigma_u <- 0.8
sigma <- 1.2

# Generate data
group <- rep(1:n_groups, each = n_per_group)
x <- rnorm(n_groups * n_per_group)
u <- rnorm(n_groups, sd = sigma_u)
y <- beta_0 + beta_1 * x + u[group] + rnorm(n_groups * n_per_group, sd = sigma)

data <- data.frame(y = y, x = x, group = group)

# View first few rows
head(data)
```

### Fit Model

```{r fit-basic}
fit <- gllamm(y ~ x + (1 | group),
              data = data,
              family = gaussian())

# Print results
print(fit)
```

### Model Summary

```{r summary-basic}
summary(fit)
```

## Extracting Components

GLLAMMR provides several functions to extract model components:

```{r extract}
# Fixed effects coefficients
fixef(fit)

# Random effects (BLUPs)
ranef(fit)

# Variance components
VarCorr(fit)

# Fitted values
head(fitted(fit))

# Residuals
head(residuals(fit))

# Log-likelihood
logLik(fit)

# AIC and BIC
AIC(fit)
BIC(fit)
```

## Random Intercept and Slope

For a more complex model with both random intercepts and slopes:

$$y_{ij} = \beta_0 + \beta_1 x_{ij} + u_{0j} + u_{1j}x_{ij} + \epsilon_{ij}$$

```{r random-slope}
# Simulate data with random slopes
set.seed(456)
n_groups <- 20
n_per_group <- 15

group <- rep(1:n_groups, each = n_per_group)
x <- rnorm(n_groups * n_per_group)

# Random intercepts and slopes
u0 <- rnorm(n_groups, sd = 1.0)
u1 <- rnorm(n_groups, sd = 0.5)

y <- 3.0 + 0.8 * x + u0[group] + u1[group] * x + rnorm(n_groups * n_per_group, sd = 1.5)

data_rs <- data.frame(y = y, x = x, group = group)

# Fit model with random slope
fit_rs <- gllamm(y ~ x + (x | group), data = data_rs)

summary(fit_rs)
```

## Three-Level Model

GLLAMMR supports models with multiple levels of nesting:

```{r three-level}
# Students nested in classes nested in schools
# y_ijk = β0 + β1*x_ijk + u_school[k] + u_class[jk] + ε_ijk

# Simulate three-level data
set.seed(789)
n_schools <- 15
n_classes_per_school <- 3
n_students_per_class <- 8

n_obs <- n_schools * n_classes_per_school * n_students_per_class

school <- rep(1:n_schools, each = n_classes_per_school * n_students_per_class)
class <- rep(1:(n_schools * n_classes_per_school), each = n_students_per_class)
x <- rnorm(n_obs)

u_school <- rnorm(n_schools, sd = 1.2)
u_class <- rnorm(n_schools * n_classes_per_school, sd = 0.8)

y <- 5.0 + 0.5 * x + u_school[school] + u_class[class] + rnorm(n_obs, sd = 1.0)

data_3level <- data.frame(y = y, x = x, school = school, class = class)

# Fit three-level model
fit_3level <- gllamm(y ~ x + (1 | school/class), data = data_3level)

summary(fit_3level)
```

## Prediction

Generate predictions from fitted models:

```{r prediction}
# Predictions with random effects (default)
pred_full <- predict(fit)

# Population-level predictions (no random effects)
pred_pop <- predict(fit, re.form = NA)

# Compare
head(data.frame(
  observed = data$y,
  fitted_with_re = pred_full,
  fitted_no_re = pred_pop
))
```

## Simulation

Simulate new data from a fitted model:

```{r simulation}
# Single simulation
sim1 <- simulate(fit, nsim = 1, seed = 999)

# Multiple simulations
sim10 <- simulate(fit, nsim = 10, seed = 999)
head(sim10)
```

## Next Steps

- See `vignette("multilevel-glmm")` for advanced GLMM features
- See `vignette("irt-models")` for item response theory models
- See `vignette("latent-class")` for latent class models
- See `vignette("mixed-response")` for models with multiple outcomes

## References

- Rabe-Hesketh, S., Skrondal, A., & Pickles, A. (2004). GLLAMM Manual.
- Skrondal, A., & Rabe-Hesketh, S. (2004). Generalized Latent Variable Modeling.
