---
title: "Item Response Theory Models with GLLAMMR"
author: "GLLAMMR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Item Response Theory Models with GLLAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## Introduction to IRT

Item Response Theory (IRT) provides a framework for analyzing test and survey data. IRT models relate person abilities (latent traits) to item responses through mathematical functions.

GLLAMMR implements the major IRT models:

- **Dichotomous**: Rasch (1PL), 2PL, 3PL
- **Polytomous**: Graded Response Model (GRM), Partial Credit Model (PCM), Generalized Partial Credit Model (GPCM), Nominal Response Model (NRM)

plus differential item functioning (DIF) analysis and explanatory IRT (EIRT) with item covariates.

Identification note: for the Rasch model the latent ability variance is estimated; for 2PL/3PL and the polytomous models the latent variance is fixed at 1 so that item discriminations are identified.

## The Rasch Model

The Rasch model is the simplest IRT model:

$$P(Y_{ij} = 1 | \theta_i, b_j) = \frac{1}{1 + \exp(-(\theta_i - b_j))}$$

where:

- $\theta_i$ is the ability of person $i$
- $b_j$ is the difficulty of item $j$

### Simulating Rasch Data

```{r rasch-simulation}
library(GLLAMMR)

set.seed(123)
n_persons <- 300
n_items <- 12

# True parameters
true_ability <- rnorm(n_persons, mean = 0, sd = 1)
true_difficulty <- seq(-2, 2, length.out = n_items)

# Generate responses
responses <- matrix(NA, n_persons, n_items)
for (j in 1:n_items) {
  p <- plogis(true_ability - true_difficulty[j])
  responses[, j] <- rbinom(n_persons, 1, p)
}
colnames(responses) <- paste0("Item", 1:n_items)

head(responses)
```

### Fitting the Rasch Model

```{r fit-rasch}
fit_rasch <- fit_irt(responses, model = "Rasch")

print(fit_rasch)
```

### Examining Item Parameters

```{r rasch-items}
# Item difficulties
item_params <- fit_rasch$item_parameters
print(item_params)

# Plot item difficulties
plot(true_difficulty, item_params$difficulty,
     xlab = "True Difficulty",
     ylab = "Estimated Difficulty",
     main = "Rasch Model: Parameter Recovery")
abline(0, 1, col = "red", lty = 2)
```

### Examining Person Parameters

```{r rasch-persons}
# Person abilities (empirical Bayes estimates)
abilities_rasch <- fit_rasch$person_abilities

# Distribution of abilities
hist(abilities_rasch, main = "Distribution of Estimated Abilities",
     xlab = "Ability (theta)", col = "lightblue")

# Correlation with true abilities
cor(abilities_rasch, true_ability)
```

## The 2PL Model

The 2-parameter logistic model adds item discrimination:

$$P(Y_{ij} = 1 | \theta_i, a_j, b_j) = \frac{1}{1 + \exp(-a_j(\theta_i - b_j))}$$

where $a_j$ is the discrimination parameter for item $j$: higher $a_j$ means the item better distinguishes between ability levels.

### Simulating 2PL Data

```{r 2pl-simulation}
set.seed(456)
n_items_2pl <- 10

true_difficulty_2pl <- rnorm(n_items_2pl, 0, 1)
true_discrimination <- runif(n_items_2pl, 0.8, 2.0)

responses_2pl <- matrix(NA, n_persons, n_items_2pl)
for (j in 1:n_items_2pl) {
  p <- plogis(true_discrimination[j] * (true_ability - true_difficulty_2pl[j]))
  responses_2pl[, j] <- rbinom(n_persons, 1, p)
}
colnames(responses_2pl) <- paste0("Q", 1:n_items_2pl)
```

### Fitting the 2PL Model

```{r fit-2pl}
fit_2pl <- fit_irt(responses_2pl, model = "2PL")

print(fit_2pl)
summary(fit_2pl)
```

### Comparing Item Parameters

```{r 2pl-params}
difficulty_2pl <- fit_2pl$item_parameters$difficulty
discrimination_2pl <- fit_2pl$item_parameters$discrimination

plot(true_difficulty_2pl, difficulty_2pl,
     xlab = "True Difficulty", ylab = "Estimated Difficulty",
     main = "2PL: Difficulty Recovery")
abline(0, 1, col = "red", lty = 2)

plot(true_discrimination, discrimination_2pl,
     xlab = "True Discrimination", ylab = "Estimated Discrimination",
     main = "2PL: Discrimination Recovery")
abline(0, 1, col = "red", lty = 2)
```

## The 3PL Model

The 3-parameter logistic model adds a guessing parameter:

$$P(Y_{ij} = 1 | \theta_i, a_j, b_j, c_j) = c_j + (1 - c_j) \frac{1}{1 + \exp(-a_j(\theta_i - b_j))}$$

where $c_j$ is the pseudo-guessing parameter (lower asymptote). Note that 3PL guessing parameters are weakly identified in small samples; large datasets are recommended in practice.

```{r fit-3pl}
set.seed(789)
true_guessing <- runif(n_items_2pl, 0.1, 0.3)

responses_3pl <- matrix(NA, n_persons, n_items_2pl)
for (j in 1:n_items_2pl) {
  p <- true_guessing[j] + (1 - true_guessing[j]) *
       plogis(true_discrimination[j] * (true_ability - true_difficulty_2pl[j]))
  responses_3pl[, j] <- rbinom(n_persons, 1, p)
}

fit_3pl <- fit_irt(responses_3pl, model = "3PL")

print(fit_3pl)
```

## Item Characteristic Curves

The `plot()` method draws item characteristic curves, item/test information, and the ability distribution; here we draw a Rasch ICC by hand:

```{r icc, fig.width=7, fig.height=6}
plot_icc_rasch <- function(difficulty, theta_range = c(-3, 3)) {
  theta <- seq(theta_range[1], theta_range[2], length.out = 100)
  p <- plogis(theta - difficulty)

  plot(theta, p, type = "l", lwd = 2,
       xlab = "Ability (theta)",
       ylab = "P(Correct)",
       main = paste("ICC: Difficulty =", round(difficulty, 2)),
       ylim = c(0, 1))
  abline(h = 0.5, col = "gray", lty = 2)
  abline(v = difficulty, col = "red", lty = 2)
}

oldpar <- par(mfrow = c(2, 2))
plot_icc_rasch(item_params$difficulty[1])
plot_icc_rasch(item_params$difficulty[4])
plot_icc_rasch(item_params$difficulty[8])
plot_icc_rasch(item_params$difficulty[12])
par(oldpar)
```

## Item Information Functions

```{r iif, fig.width=6, fig.height=4.5}
item_information_2pl <- function(theta, a, b) {
  p <- plogis(a * (theta - b))
  a^2 * p * (1 - p)
}

theta_grid <- seq(-3, 3, length.out = 100)

plot(theta_grid, item_information_2pl(theta_grid, a = 0.5, b = 0),
     type = "l", lwd = 2, col = "blue",
     xlab = "Ability", ylab = "Information",
     main = "Item Information: Effect of Discrimination",
     ylim = c(0, 1))
lines(theta_grid, item_information_2pl(theta_grid, a = 1.0, b = 0),
      lwd = 2, col = "green")
lines(theta_grid, item_information_2pl(theta_grid, a = 2.0, b = 0),
      lwd = 2, col = "red")
legend("topright",
       legend = c("a = 0.5", "a = 1.0", "a = 2.0"),
       col = c("blue", "green", "red"),
       lwd = 2)
```

## Model Comparison

Models fitted to the *same* data can be compared with information criteria. Here we fit both Rasch and 2PL to the 2PL data:

```{r model-comparison}
fit_rasch_on_2pl <- fit_irt(responses_2pl, model = "Rasch")

comparison <- data.frame(
  Model = c("Rasch", "2PL"),
  LogLik = c(fit_rasch_on_2pl$logLik, fit_2pl$logLik),
  AIC = c(fit_rasch_on_2pl$AIC, fit_2pl$AIC),
  BIC = c(fit_rasch_on_2pl$BIC, fit_2pl$BIC)
)

print(comparison)
# Lower AIC/BIC is better; 2PL should fit better here because the
# generating discriminations vary across items
```

## Ability Estimation

```{r ability-est}
# Create ability score groups
ability_groups <- cut(abilities_rasch,
                      breaks = quantile(abilities_rasch, c(0, 0.25, 0.5, 0.75, 1)),
                      include.lowest = TRUE,
                      labels = c("Low", "Medium-Low", "Medium-High", "High"))

# Average proportion correct by ability group
score_by_group <- tapply(rowMeans(responses), ability_groups, mean)
print(score_by_group)
```

## Polytomous IRT Models

### Graded Response Model (GRM)

The Graded Response Model is used for ordered categorical responses (e.g., Likert scales):

$$P(Y_{ij} \geq k | \theta_i, a_j, b_{jk}) = \frac{1}{1 + \exp(-a_j(\theta_i - b_{jk}))}$$

where $b_{j1} < b_{j2} < ... < b_{j,K-1}$ are ordered thresholds.

#### Simulating GRM Data

```{r grm-simulation}
set.seed(123)
n_persons_grm <- 200
n_items_grm <- 8
n_categories <- 4

theta_grm <- rnorm(n_persons_grm, 0, 1)
discrimination_grm <- runif(n_items_grm, 0.8, 2.0)
difficulty_base <- rnorm(n_items_grm, 0, 1)

responses_grm <- matrix(NA, n_persons_grm, n_items_grm)
for (j in 1:n_items_grm) {
  # Ordered thresholds centered around the item difficulty
  thresholds <- difficulty_base[j] + seq(-1.5, 1.5, length.out = n_categories - 1)

  for (i in 1:n_persons_grm) {
    # Cumulative probabilities P(Y >= k), k = 2..K
    cum <- plogis(discrimination_grm[j] * (theta_grm[i] - thresholds))
    # Category probabilities P(Y = k)
    probs <- c(1 - cum[1], -diff(cum), cum[n_categories - 1])
    responses_grm[i, j] <- sample(1:n_categories, 1, prob = probs)
  }
}

# View response distribution for the first item
table(responses_grm[, 1])
```

#### Fitting the GRM

```{r fit-grm}
fit_grm <- fit_irt(responses_grm, model = "GRM")

print(fit_grm)
summary(fit_grm)
```

#### Examining GRM Parameters

```{r grm-params}
# Item discriminations
discriminations <- fit_grm$item_parameters$discrimination
print(head(discriminations))

# Thresholds (one per category transition)
thresholds <- fit_grm$item_parameters$thresholds
print(thresholds[[1]])  # Thresholds for first item
```

### Partial Credit Model (PCM)

The Partial Credit Model is used for partial credit scoring:

$$P(Y_{ij} = k | \theta_i, \delta_{jk}) = \frac{\exp(\sum_{h=0}^{k}(\theta_i - \delta_{jh}))}{\sum_{m=0}^{K}\exp(\sum_{h=0}^{m}(\theta_i - \delta_{jh}))}$$

```{r fit-pcm}
# PCM constrains discrimination = 1 for all items
fit_pcm <- fit_irt(responses_grm, model = "PCM")

print(fit_pcm)
```

### Generalized Partial Credit Model (GPCM)

GPCM extends PCM by allowing item-specific discriminations:

```{r fit-gpcm}
fit_gpcm <- fit_irt(responses_grm, model = "GPCM")

print(fit_gpcm)
```

### Nominal Response Model (NRM)

NRM is for unordered categorical responses (no natural ordering):

```{r nrm-simulation}
# Simulate nominal responses (e.g., multiple choice with plausible distractors)
set.seed(456)
responses_nrm <- matrix(sample(1:4, 200 * 8, replace = TRUE), 200, 8)

fit_nrm <- fit_irt(responses_nrm, model = "NRM")
print(fit_nrm)
```

### Comparing Polytomous Models

```{r compare-polytomous}
comparison_poly <- data.frame(
  Model = c("GRM", "PCM", "GPCM"),
  LogLik = c(fit_grm$logLik, fit_pcm$logLik, fit_gpcm$logLik),
  AIC = c(fit_grm$AIC, fit_pcm$AIC, fit_gpcm$AIC),
  BIC = c(fit_grm$BIC, fit_pcm$BIC, fit_gpcm$BIC)
)

print(comparison_poly)
# GRM/GPCM typically fit better than PCM if discriminations vary;
# PCM is more parsimonious (fewer parameters)
```

## Differential Item Functioning (DIF)

DIF analysis tests whether items function differently across groups (e.g., gender, ethnicity, language).

- **Uniform DIF**: item difficulty differs between groups (parallel ICCs)
- **Non-uniform DIF**: item discrimination differs between groups (non-parallel ICCs)

### Simulating Data with DIF

```{r dif-simulation}
set.seed(789)
n_persons_dif <- 300
n_items_dif <- 12

group <- rep(c("Group A", "Group B"), each = n_persons_dif / 2)
theta_dif <- rnorm(n_persons_dif, 0, 1)

difficulty_dif <- rnorm(n_items_dif, 0, 1)
discrimination_dif <- runif(n_items_dif, 0.8, 2.0)

# Add uniform DIF to items 3, 6, and 9 for Group B
dif_items <- c(3, 6, 9)
dif_amount <- c(0.8, 1.2, -0.6)

responses_dif <- matrix(NA, n_persons_dif, n_items_dif)
for (i in 1:n_persons_dif) {
  for (j in 1:n_items_dif) {
    diff_effective <- difficulty_dif[j]
    if (j %in% dif_items && group[i] == "Group B") {
      diff_effective <- diff_effective + dif_amount[which(dif_items == j)]
    }
    p <- plogis(discrimination_dif[j] * (theta_dif[i] - diff_effective))
    responses_dif[i, j] <- rbinom(1, 1, p)
  }
}
```

### Testing for DIF

```{r dif-test}
dif_result <- dif_test(
  responses_dif,
  dif = group,
  model = "2PL",
  purify = TRUE,
  alpha = 0.05
)

print(dif_result)
```

`dif_test()` uses logistic-regression DIF with a latent (EAP) matching
criterion and iterative purification: items flagged in one round are
removed from the matching criterion for the next, so DIF items do not
contaminate the score against which DIF is judged. Multiple DIF
variables and their interactions are specified with a formula, e.g.
`dif = ~ gender * language` with `person_data`.

### Interpreting DIF Results

```{r dif-interpret}
summary(dif_result)

# Items flagged for DIF
cat("Flagged items:", dif_result$flagged_items, "\n")

# Effect sizes for flagged items
dif_results_df <- dif_result$dif_results
flagged_df <- dif_results_df[dif_results_df$item %in% dif_result$flagged_items, ]
print(flagged_df)

# Effect sizes are Nagelkerke delta-R2 between the compared models
# (Jodoin-Gierl): A < 0.035 negligible, B < 0.07 moderate, C large
```

### Visualizing DIF

```{r dif-plot}
# Plot ICCs by group for an item with simulated DIF
dif_plot(dif_result, item = 3)
```

### DIF with Polytomous Items

```{r dif-polytomous}
# DIF analysis also works with polytomous models
dif_result_grm <- dif_test(
  responses_grm,
  dif = sample(c("Male", "Female"), nrow(responses_grm), replace = TRUE),
  model = "GRM",
  alpha = 0.05
)

print(dif_result_grm)
```

## Explanatory IRT (EIRT)

Explanatory IRT models item parameters as functions of item characteristics (covariates). Instead of estimating $b_j$ and $a_j$ separately for each item, model them as:

$$b_j = \gamma_0 + \gamma_1 W_{j1} + \gamma_2 W_{j2} + ... + \epsilon_{bj}$$
$$\log(a_j) = \delta_0 + \delta_1 V_{j1} + \delta_2 V_{j2} + ... + \epsilon_{aj}$$

where $W$ and $V$ are item-level covariates.

### Example: Reading Test Items

```{r eirt-setup}
set.seed(111)
n_persons_eirt <- 300
n_items_eirt <- 15

# Item characteristics
item_data <- data.frame(
  word_frequency = rnorm(n_items_eirt, 0, 1),  # Higher = more common words
  passage_length = rpois(n_items_eirt, 100),   # Length in words
  item_type = factor(sample(c("Literal", "Inferential", "Critical"),
                            n_items_eirt, replace = TRUE))
)

# True covariate effects
theta_eirt <- rnorm(n_persons_eirt, 0, 1)

difficulty_eirt <- -0.6 * item_data$word_frequency +
                    0.003 * item_data$passage_length +
                    rnorm(n_items_eirt, 0, 0.3)

discrimination_eirt <- exp(0.3 +
                           0.4 * (item_data$item_type == "Inferential") +
                           0.6 * (item_data$item_type == "Critical") +
                           rnorm(n_items_eirt, 0, 0.2))

responses_eirt <- matrix(NA, n_persons_eirt, n_items_eirt)
for (j in 1:n_items_eirt) {
  p <- plogis(discrimination_eirt[j] * (theta_eirt - difficulty_eirt[j]))
  responses_eirt[, j] <- rbinom(n_persons_eirt, 1, p)
}
```

### Fitting EIRT Models

```{r fit-eirt}
fit_eirt_model <- fit_eirt(
  responses_eirt,
  item_data = item_data,
  difficulty_formula = ~ word_frequency + passage_length,
  discrimination_formula = ~ item_type,
  model = "2PL"
)

print(fit_eirt_model)
```

### Interpreting EIRT Results

```{r eirt-results}
# Difficulty regression coefficients
gamma_hat <- fit_eirt_model$regression_coefficients$difficulty
print(gamma_hat)
# Negative word_frequency: items with common words are easier
# Positive passage_length: longer passages are harder

# Discrimination regression coefficients (log scale)
delta_hat <- fit_eirt_model$regression_coefficients$discrimination
print(delta_hat)
# Positive coefficients for Inferential/Critical: those item types
# discriminate better than Literal items

# Residual standard deviations: variation remaining after covariates
print(fit_eirt_model$residual_sd)
```

### Comparing EIRT to Standard IRT

```{r eirt-comparison}
fit_standard <- fit_irt(responses_eirt, model = "2PL")

cat("Standard IRT:    logLik =", round(fit_standard$logLik, 1),
    " AIC =", round(fit_standard$AIC, 1), "\n")
cat("Explanatory IRT: logLik =", round(fit_eirt_model$logLik, 1),
    " AIC =", round(fit_eirt_model$AIC, 1), "\n")

# EIRT can have better AIC when covariates explain substantial variation,
# and always provides interpretable regression coefficients
```

### EIRT with Polytomous Items

```{r eirt-polytomous}
# Item characteristics for the Likert-scale (GRM) items
item_data_likert <- data.frame(
  complexity = rnorm(n_items_grm, 0, 1),
  reverse_coded = factor(sample(c("Yes", "No"), n_items_grm, replace = TRUE))
)

fit_eirt_grm <- fit_eirt(
  responses_grm,
  item_data = item_data_likert,
  difficulty_formula = ~ complexity + reverse_coded,
  discrimination_formula = ~ complexity,
  model = "GRM"
)

print(fit_eirt_grm)
```

### Practical Applications of EIRT

Use EIRT to identify item characteristics that predict difficulty, design new items with targeted difficulty, and balance test forms. For example, predicting the difficulty of a not-yet-written item from its planned characteristics:

```{r test-development}
new_item <- data.frame(
  word_frequency = 0.5,  # Moderately common words
  passage_length = 90    # Short passage
)

W_new <- model.matrix(~ word_frequency + passage_length, data = new_item)
predicted_difficulty <- as.numeric(W_new %*% gamma_hat)
cat("Predicted difficulty for new item:", round(predicted_difficulty, 2), "\n")
```

Items whose estimated parameters deviate strongly from their covariate-based predictions are candidates for review:

```{r item-revision}
predicted_diffs <- model.matrix(~ word_frequency + passage_length,
                                data = item_data) %*% gamma_hat
residuals_diff <- fit_eirt_model$item_parameters$difficulty - predicted_diffs
print(round(as.numeric(residuals_diff), 2))
```

## Multilevel IRT

Person-level grouping structures (students in classes in schools) are supported via the `random` argument of `fit_irt()`; see `vignette("multilevel-irt")`.

## References

- Rasch, G. (1960). Probabilistic Models for Some Intelligence and Attainment Tests.
- Lord, F. M. (1980). Applications of Item Response Theory to Practical Testing Problems.
- Embretson, S. E., & Reise, S. P. (2000). Item Response Theory for Psychologists.
- De Boeck, P., & Wilson, M. (2004). Explanatory Item Response Models.

## See Also

- `vignette("multilevel-irt")` - Multilevel IRT models
- `vignette("latent-class")` - Latent class models
- `vignette("marginal-predictions")` - Population-averaged predictions
