---
title: "Item Response Theory Models with GLLAMMR"
author: "GLLAMMR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Item Response Theory Models with GLLAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction to IRT

Item Response Theory (IRT) provides a framework for analyzing test and survey data. IRT models relate person abilities (latent traits) to item responses through mathematical functions.

GLLAMMR implements major IRT models:
- **Rasch model** (1-parameter logistic)
- **2PL model** (2-parameter logistic)
- **3PL model** (3-parameter logistic with guessing)

## The Rasch Model

The Rasch model is the simplest IRT model:

$$P(Y_{ij} = 1 | \theta_i, b_j) = \frac{1}{1 + \exp(-([\theta_i - b_j]))}$$

where:
- $\theta_i$ is the ability of person $i$
- $b_j$ is the difficulty of item $j$

### Simulating Rasch Data

```{r rasch-simulation}
library(GLLAMMR)

set.seed(123)
n_persons <- 500
n_items <- 20

# True parameters
true_ability <- rnorm(n_persons, mean = 0, sd = 1)
true_difficulty <- seq(-2, 2, length.out = n_items)

# Generate responses
responses <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    p <- plogis(true_ability[i] - true_difficulty[j])
    responses[i, j] <- rbinom(1, 1, p)
  }
}

# Item names
colnames(responses) <- paste0("Item", 1:n_items)

# View first few rows
head(responses)
```

### Fitting the Rasch Model

```{r fit-rasch}
fit_rasch <- fit_irt(responses, model = "Rasch")

print(fit_rasch)
```

### Examining Item Parameters

```{r rasch-items}
# Item difficulties
item_params <- fit_rasch$item_parameters
print(item_params)

# Plot item difficulties
plot(true_difficulty, item_params$difficulty,
     xlab = "True Difficulty",
     ylab = "Estimated Difficulty",
     main = "Rasch Model: Parameter Recovery")
abline(0, 1, col = "red", lty = 2)
```

### Examining Person Parameters

```{r rasch-persons}
# Person abilities
abilities <- fit_rasch$person_abilities

# Distribution of abilities
hist(abilities, main = "Distribution of Estimated Abilities",
     xlab = "Ability (theta)", col = "lightblue")

# Correlation with true abilities
cor(abilities, true_ability)
```

## The 2PL Model

The 2-parameter logistic model adds item discrimination:

$$P(Y_{ij} = 1 | \theta_i, a_j, b_j) = \frac{1}{1 + \exp(-a_j(\theta_i - b_j))}$$

where:
- $a_j$ is the discrimination parameter for item $j$
- Higher $a_j$ means the item better distinguishes between ability levels

### Simulating 2PL Data

```{r 2pl-simulation}
set.seed(456)
n_persons <- 500
n_items <- 15

true_ability <- rnorm(n_persons, 0, 1)
true_difficulty <- rnorm(n_items, 0, 1)
true_discrimination <- runif(n_items, 0.5, 2.5)

responses_2pl <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    p <- plogis(true_discrimination[j] * (true_ability[i] - true_difficulty[j]))
    responses_2pl[i, j] <- rbinom(1, 1, p)
  }
}

colnames(responses_2pl) <- paste0("Q", 1:n_items)
```

### Fitting the 2PL Model

```{r fit-2pl}
fit_2pl <- fit_irt(responses_2pl, model = "2PL")

print(fit_2pl)
summary(fit_2pl)
```

### Comparing Item Parameters

```{r 2pl-params}
# Extract parameters
difficulty_2pl <- fit_2pl$item_parameters$difficulty
discrimination_2pl <- fit_2pl$item_parameters$discrimination

# Plot difficulties
plot(true_difficulty, difficulty_2pl,
     xlab = "True Difficulty", ylab = "Estimated Difficulty",
     main = "2PL: Difficulty Recovery")
abline(0, 1, col = "red", lty = 2)

# Plot discriminations
plot(true_discrimination, discrimination_2pl,
     xlab = "True Discrimination", ylab = "Estimated Discrimination",
     main = "2PL: Discrimination Recovery")
abline(0, 1, col = "red", lty = 2)
```

## The 3PL Model

The 3-parameter logistic model adds a guessing parameter:

$$P(Y_{ij} = 1 | \theta_i, a_j, b_j, c_j) = c_j + (1 - c_j) \frac{1}{1 + \exp(-a_j(\theta_i - b_j))}$$

where $c_j$ is the pseudo-guessing parameter (lower asymptote).

### Fitting the 3PL Model

```{r fit-3pl}
# Add guessing to data generation
set.seed(789)
true_guessing <- runif(n_items, 0.1, 0.3)

responses_3pl <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    p <- true_guessing[j] + (1 - true_guessing[j]) *
         plogis(true_discrimination[j] * (true_ability[i] - true_difficulty[j]))
    responses_3pl[i, j] <- rbinom(1, 1, p)
  }
}

fit_3pl <- fit_irt(responses_3pl, model = "3PL")

print(fit_3pl)
```

## Item Characteristic Curves

```{r icc, eval=FALSE}
# Function to plot ICC for Rasch model
plot_icc_rasch <- function(difficulty, theta_range = c(-3, 3)) {
  theta <- seq(theta_range[1], theta_range[2], length.out = 100)
  p <- plogis(theta - difficulty)

  plot(theta, p, type = "l", lwd = 2,
       xlab = "Ability (theta)",
       ylab = "P(Correct)",
       main = paste("ICC: Difficulty =", round(difficulty, 2)),
       ylim = c(0, 1))
  abline(h = 0.5, col = "gray", lty = 2)
  abline(v = difficulty, col = "red", lty = 2)
}

# Plot ICCs for selected items
par(mfrow = c(2, 2))
plot_icc_rasch(item_params$difficulty[1])
plot_icc_rasch(item_params$difficulty[5])
plot_icc_rasch(item_params$difficulty[10])
plot_icc_rasch(item_params$difficulty[15])
par(mfrow = c(1, 1))
```

## Item Information Functions

```{r iif, eval=FALSE}
# Function to compute item information for 2PL
item_information_2pl <- function(theta, a, b) {
  p <- plogis(a * (theta - b))
  info <- a^2 * p * (1 - p)
  return(info)
}

# Plot information for items with different discriminations
theta <- seq(-3, 3, length.out = 100)

plot(theta, item_information_2pl(theta, a = 0.5, b = 0),
     type = "l", lwd = 2, col = "blue",
     xlab = "Ability", ylab = "Information",
     main = "Item Information: Effect of Discrimination",
     ylim = c(0, 1))
lines(theta, item_information_2pl(theta, a = 1.0, b = 0),
      lwd = 2, col = "green")
lines(theta, item_information_2pl(theta, a = 2.0, b = 0),
      lwd = 2, col = "red")
legend("topright",
       legend = c("a = 0.5", "a = 1.0", "a = 2.0"),
       col = c("blue", "green", "red"),
       lwd = 2)
```

## Model Comparison

```{r model-comparison, eval=FALSE}
# Compare Rasch vs 2PL
comparison <- data.frame(
  Model = c("Rasch", "2PL", "3PL"),
  LogLik = c(fit_rasch$logLik, fit_2pl$logLik, fit_3pl$logLik),
  AIC = c(fit_rasch$AIC, fit_2pl$AIC, fit_3pl$AIC),
  BIC = c(fit_rasch$BIC, fit_2pl$BIC, fit_3pl$BIC)
)

print(comparison)

# Lower AIC/BIC is better
# 2PL should fit better than Rasch if discrimination varies
# 3PL should fit better if there's guessing
```

## Ability Estimation

```{r ability-est}
# Get person abilities from fitted model
abilities_rasch <- fit_rasch$person_abilities

# Create ability score groups
ability_groups <- cut(abilities_rasch,
                      breaks = quantile(abilities_rasch, c(0, 0.25, 0.5, 0.75, 1)),
                      include.lowest = TRUE,
                      labels = c("Low", "Medium-Low", "Medium-High", "High"))

# Average score by ability group
score_by_group <- tapply(rowMeans(responses), ability_groups, mean)
print(score_by_group)
```

## Practical Applications

### Educational Assessment

```{r education, eval=FALSE}
# Simulate reading test data
n_students <- 1000
n_questions <- 30

# Items vary in difficulty (easy to hard)
difficulty_reading <- seq(-2, 2, length.out = n_questions)

# Generate abilities (normally distributed)
student_ability <- rnorm(n_students, mean = 0, sd = 1)

# Generate responses
reading_test <- matrix(NA, n_students, n_questions)
for (i in 1:n_students) {
  for (j in 1:n_questions) {
    p <- plogis(student_ability[i] - difficulty_reading[j])
    reading_test[i, j] <- rbinom(1, 1, p)
  }
}

# Fit model
fit_reading <- fit_irt(reading_test, model = "Rasch")

# Identify problematic items
item_fit <- fit_reading$item_parameters

# Items that are too easy or too hard
extreme_items <- which(abs(item_fit$difficulty) > 2)
print(paste("Extreme items:", paste(extreme_items, collapse = ", ")))
```

### Psychological Scales

```{r psychology, eval=FALSE}
# Simulate depression scale
n_respondents <- 800
n_symptoms <- 15

# Symptoms have varying severity (difficulty)
symptom_severity <- rnorm(n_symptoms, mean = 0, sd = 1)

# Respondents have varying depression levels
depression_level <- rnorm(n_respondents, mean = 0, sd = 1.5)

# Generate endorsements
depression_scale <- matrix(NA, n_respondents, n_symptoms)
for (i in 1:n_respondents) {
  for (j in 1:n_symptoms) {
    p <- plogis(depression_level[i] - symptom_severity[j])
    depression_scale[i, j] <- rbinom(1, 1, p)
  }
}

# Fit model
fit_depression <- fit_irt(depression_scale, model = "Rasch")

# Score respondents
scores <- fit_depression$person_abilities

# Identify high-risk individuals (top 10%)
high_risk_cutoff <- quantile(scores, 0.90)
high_risk <- which(scores > high_risk_cutoff)
print(paste("Number of high-risk individuals:", length(high_risk)))
```

## Advanced Topics

### Differential Item Functioning (DIF)

```{r dif, eval=FALSE}
# Future feature: Test if items function differently across groups
# fit_dif <- fit_irt_dif(responses, group = gender, model = "Rasch")
```

### Multidimensional IRT

```{r mirt, eval=FALSE}
# Future feature: Multiple latent dimensions
# fit_mirt <- fit_irt(responses, model = "2PL", dimensions = 2)
```

## References

- Rasch, G. (1960). Probabilistic Models for Some Intelligence and Attainment Tests.
- Lord, F. M. (1980). Applications of Item Response Theory to Practical Testing Problems.
- Embretson, S. E., & Reise, S. P. (2000). Item Response Theory for Psychologists.
- Baker, F. B., & Kim, S. H. (2004). Item Response Theory: Parameter Estimation Techniques.

## See Also

- `vignette("latent-class")` - Latent class models
- `vignette("factor-analysis")` - Factor analysis with GLLAMMR
- `vignette("mixed-response")` - Mixed response models
