---
title: "Item Response Theory Models with GLLAMMR"
author: "GLLAMMR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Item Response Theory Models with GLLAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction to IRT

Item Response Theory (IRT) provides a framework for analyzing test and survey data. IRT models relate person abilities (latent traits) to item responses through mathematical functions.

GLLAMMR implements major IRT models:
- **Rasch model** (1-parameter logistic)
- **2PL model** (2-parameter logistic)
- **3PL model** (3-parameter logistic with guessing)

## The Rasch Model

The Rasch model is the simplest IRT model:

$$P(Y_{ij} = 1 | \theta_i, b_j) = \frac{1}{1 + \exp(-([\theta_i - b_j]))}$$

where:
- $\theta_i$ is the ability of person $i$
- $b_j$ is the difficulty of item $j$

### Simulating Rasch Data

```{r rasch-simulation}
library(GLLAMMR)

set.seed(123)
n_persons <- 500
n_items <- 20

# True parameters
true_ability <- rnorm(n_persons, mean = 0, sd = 1)
true_difficulty <- seq(-2, 2, length.out = n_items)

# Generate responses
responses <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    p <- plogis(true_ability[i] - true_difficulty[j])
    responses[i, j] <- rbinom(1, 1, p)
  }
}

# Item names
colnames(responses) <- paste0("Item", 1:n_items)

# View first few rows
head(responses)
```

### Fitting the Rasch Model

```{r fit-rasch}
fit_rasch <- fit_irt(responses, model = "Rasch")

print(fit_rasch)
```

### Examining Item Parameters

```{r rasch-items}
# Item difficulties
item_params <- fit_rasch$item_parameters
print(item_params)

# Plot item difficulties
plot(true_difficulty, item_params$difficulty,
     xlab = "True Difficulty",
     ylab = "Estimated Difficulty",
     main = "Rasch Model: Parameter Recovery")
abline(0, 1, col = "red", lty = 2)
```

### Examining Person Parameters

```{r rasch-persons}
# Person abilities
abilities <- fit_rasch$person_abilities

# Distribution of abilities
hist(abilities, main = "Distribution of Estimated Abilities",
     xlab = "Ability (theta)", col = "lightblue")

# Correlation with true abilities
cor(abilities, true_ability)
```

## The 2PL Model

The 2-parameter logistic model adds item discrimination:

$$P(Y_{ij} = 1 | \theta_i, a_j, b_j) = \frac{1}{1 + \exp(-a_j(\theta_i - b_j))}$$

where:
- $a_j$ is the discrimination parameter for item $j$
- Higher $a_j$ means the item better distinguishes between ability levels

### Simulating 2PL Data

```{r 2pl-simulation}
set.seed(456)
n_persons <- 500
n_items <- 15

true_ability <- rnorm(n_persons, 0, 1)
true_difficulty <- rnorm(n_items, 0, 1)
true_discrimination <- runif(n_items, 0.5, 2.5)

responses_2pl <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    p <- plogis(true_discrimination[j] * (true_ability[i] - true_difficulty[j]))
    responses_2pl[i, j] <- rbinom(1, 1, p)
  }
}

colnames(responses_2pl) <- paste0("Q", 1:n_items)
```

### Fitting the 2PL Model

```{r fit-2pl}
fit_2pl <- fit_irt(responses_2pl, model = "2PL")

print(fit_2pl)
summary(fit_2pl)
```

### Comparing Item Parameters

```{r 2pl-params}
# Extract parameters
difficulty_2pl <- fit_2pl$item_parameters$difficulty
discrimination_2pl <- fit_2pl$item_parameters$discrimination

# Plot difficulties
plot(true_difficulty, difficulty_2pl,
     xlab = "True Difficulty", ylab = "Estimated Difficulty",
     main = "2PL: Difficulty Recovery")
abline(0, 1, col = "red", lty = 2)

# Plot discriminations
plot(true_discrimination, discrimination_2pl,
     xlab = "True Discrimination", ylab = "Estimated Discrimination",
     main = "2PL: Discrimination Recovery")
abline(0, 1, col = "red", lty = 2)
```

## The 3PL Model

The 3-parameter logistic model adds a guessing parameter:

$$P(Y_{ij} = 1 | \theta_i, a_j, b_j, c_j) = c_j + (1 - c_j) \frac{1}{1 + \exp(-a_j(\theta_i - b_j))}$$

where $c_j$ is the pseudo-guessing parameter (lower asymptote).

### Fitting the 3PL Model

```{r fit-3pl}
# Add guessing to data generation
set.seed(789)
true_guessing <- runif(n_items, 0.1, 0.3)

responses_3pl <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    p <- true_guessing[j] + (1 - true_guessing[j]) *
         plogis(true_discrimination[j] * (true_ability[i] - true_difficulty[j]))
    responses_3pl[i, j] <- rbinom(1, 1, p)
  }
}

fit_3pl <- fit_irt(responses_3pl, model = "3PL")

print(fit_3pl)
```

## Item Characteristic Curves

```{r icc, eval=FALSE}
# Function to plot ICC for Rasch model
plot_icc_rasch <- function(difficulty, theta_range = c(-3, 3)) {
  theta <- seq(theta_range[1], theta_range[2], length.out = 100)
  p <- plogis(theta - difficulty)

  plot(theta, p, type = "l", lwd = 2,
       xlab = "Ability (theta)",
       ylab = "P(Correct)",
       main = paste("ICC: Difficulty =", round(difficulty, 2)),
       ylim = c(0, 1))
  abline(h = 0.5, col = "gray", lty = 2)
  abline(v = difficulty, col = "red", lty = 2)
}

# Plot ICCs for selected items
par(mfrow = c(2, 2))
plot_icc_rasch(item_params$difficulty[1])
plot_icc_rasch(item_params$difficulty[5])
plot_icc_rasch(item_params$difficulty[10])
plot_icc_rasch(item_params$difficulty[15])
par(mfrow = c(1, 1))
```

## Item Information Functions

```{r iif, eval=FALSE}
# Function to compute item information for 2PL
item_information_2pl <- function(theta, a, b) {
  p <- plogis(a * (theta - b))
  info <- a^2 * p * (1 - p)
  return(info)
}

# Plot information for items with different discriminations
theta <- seq(-3, 3, length.out = 100)

plot(theta, item_information_2pl(theta, a = 0.5, b = 0),
     type = "l", lwd = 2, col = "blue",
     xlab = "Ability", ylab = "Information",
     main = "Item Information: Effect of Discrimination",
     ylim = c(0, 1))
lines(theta, item_information_2pl(theta, a = 1.0, b = 0),
      lwd = 2, col = "green")
lines(theta, item_information_2pl(theta, a = 2.0, b = 0),
      lwd = 2, col = "red")
legend("topright",
       legend = c("a = 0.5", "a = 1.0", "a = 2.0"),
       col = c("blue", "green", "red"),
       lwd = 2)
```

## Model Comparison

```{r model-comparison, eval=FALSE}
# Compare Rasch vs 2PL
comparison <- data.frame(
  Model = c("Rasch", "2PL", "3PL"),
  LogLik = c(fit_rasch$logLik, fit_2pl$logLik, fit_3pl$logLik),
  AIC = c(fit_rasch$AIC, fit_2pl$AIC, fit_3pl$AIC),
  BIC = c(fit_rasch$BIC, fit_2pl$BIC, fit_3pl$BIC)
)

print(comparison)

# Lower AIC/BIC is better
# 2PL should fit better than Rasch if discrimination varies
# 3PL should fit better if there's guessing
```

## Ability Estimation

```{r ability-est}
# Get person abilities from fitted model
abilities_rasch <- fit_rasch$person_abilities

# Create ability score groups
ability_groups <- cut(abilities_rasch,
                      breaks = quantile(abilities_rasch, c(0, 0.25, 0.5, 0.75, 1)),
                      include.lowest = TRUE,
                      labels = c("Low", "Medium-Low", "Medium-High", "High"))

# Average score by ability group
score_by_group <- tapply(rowMeans(responses), ability_groups, mean)
print(score_by_group)
```

## Practical Applications

### Educational Assessment

```{r education, eval=FALSE}
# Simulate reading test data
n_students <- 1000
n_questions <- 30

# Items vary in difficulty (easy to hard)
difficulty_reading <- seq(-2, 2, length.out = n_questions)

# Generate abilities (normally distributed)
student_ability <- rnorm(n_students, mean = 0, sd = 1)

# Generate responses
reading_test <- matrix(NA, n_students, n_questions)
for (i in 1:n_students) {
  for (j in 1:n_questions) {
    p <- plogis(student_ability[i] - difficulty_reading[j])
    reading_test[i, j] <- rbinom(1, 1, p)
  }
}

# Fit model
fit_reading <- fit_irt(reading_test, model = "Rasch")

# Identify problematic items
item_fit <- fit_reading$item_parameters

# Items that are too easy or too hard
extreme_items <- which(abs(item_fit$difficulty) > 2)
print(paste("Extreme items:", paste(extreme_items, collapse = ", ")))
```

### Psychological Scales

```{r psychology, eval=FALSE}
# Simulate depression scale
n_respondents <- 800
n_symptoms <- 15

# Symptoms have varying severity (difficulty)
symptom_severity <- rnorm(n_symptoms, mean = 0, sd = 1)

# Respondents have varying depression levels
depression_level <- rnorm(n_respondents, mean = 0, sd = 1.5)

# Generate endorsements
depression_scale <- matrix(NA, n_respondents, n_symptoms)
for (i in 1:n_respondents) {
  for (j in 1:n_symptoms) {
    p <- plogis(depression_level[i] - symptom_severity[j])
    depression_scale[i, j] <- rbinom(1, 1, p)
  }
}

# Fit model
fit_depression <- fit_irt(depression_scale, model = "Rasch")

# Score respondents
scores <- fit_depression$person_abilities

# Identify high-risk individuals (top 10%)
high_risk_cutoff <- quantile(scores, 0.90)
high_risk <- which(scores > high_risk_cutoff)
print(paste("Number of high-risk individuals:", length(high_risk)))
```

## Polytomous IRT Models

### Graded Response Model (GRM)

The Graded Response Model is used for ordered categorical responses (e.g., Likert scales):

$$P(Y_{ij} \geq k | \theta_i, a_j, b_{jk}) = \frac{1}{1 + \exp(-a_j(\theta_i - b_{jk}))}$$

where $b_{j1} < b_{j2} < ... < b_{j,K-1}$ are ordered thresholds.

#### Simulating GRM Data

```{r grm-simulation}
set.seed(123)
n_persons <- 300
n_items <- 15
n_categories <- 5  # e.g., 1="Strongly Disagree" to 5="Strongly Agree"

# Generate person abilities
theta <- rnorm(n_persons, 0, 1)

# Generate item parameters
discrimination <- runif(n_items, 0.8, 2.5)
difficulty_base <- rnorm(n_items, 0, 1)

# Create ordered thresholds for each item
responses_grm <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    # Thresholds centered around item difficulty
    thresholds <- difficulty_base[j] + seq(-2, 2, length.out = n_categories - 1)

    # Compute cumulative probabilities
    probs <- numeric(n_categories)
    probs[1] <- plogis(discrimination[j] * (theta[i] - thresholds[1]))

    for (k in 2:(n_categories - 1)) {
      p_le_k <- plogis(discrimination[j] * (theta[i] - thresholds[k]))
      p_le_k_minus_1 <- plogis(discrimination[j] * (theta[i] - thresholds[k-1]))
      probs[k] <- p_le_k - p_le_k_minus_1
    }

    probs[n_categories] <- 1 - plogis(discrimination[j] * (theta[i] - thresholds[n_categories-1]))

    # Sample response
    responses_grm[i, j] <- sample(1:n_categories, 1, prob = probs)
  }
}

# View response distribution
table(responses_grm[, 1])
```

#### Fitting the GRM

```{r fit-grm}
fit_grm <- fit_irt(responses_grm, model = "GRM")

print(fit_grm)
summary(fit_grm)
```

#### Examining GRM Parameters

```{r grm-params}
# Item discriminations
discriminations <- fit_grm$item_parameters$discrimination
print(head(discriminations))

# Thresholds (one per category transition)
thresholds <- fit_grm$item_parameters$thresholds
print(thresholds[[1]])  # Thresholds for first item
```

### Partial Credit Model (PCM)

The Partial Credit Model is used for partial credit scoring:

$$P(Y_{ij} = k | \theta_i, \delta_{jk}) = \frac{\exp(\sum_{h=0}^{k}(\theta_i - \delta_{jh}))}{\sum_{m=0}^{K}\exp(\sum_{h=0}^{m}(\theta_i - \delta_{jh}))}$$

#### Fitting the PCM

```{r fit-pcm}
# PCM constrains discrimination = 1 for all items
fit_pcm <- fit_irt(responses_grm, model = "PCM")

print(fit_pcm)
```

### Generalized Partial Credit Model (GPCM)

GPCM extends PCM by allowing item-specific discriminations:

```{r fit-gpcm}
fit_gpcm <- fit_irt(responses_grm, model = "GPCM")

print(fit_gpcm)
```

### Nominal Response Model (NRM)

NRM is for unordered categorical responses (no natural ordering):

```{r nrm-simulation}
# Simulate nominal responses (e.g., multiple choice with plausible distractors)
set.seed(456)
n_persons <- 250
n_items <- 12
n_options <- 4

responses_nrm <- matrix(sample(1:n_options, n_persons * n_items, replace = TRUE),
                        n_persons, n_items)

fit_nrm <- fit_irt(responses_nrm, model = "NRM")
print(fit_nrm)
```

### Comparing Polytomous Models

```{r compare-polytomous}
# Model comparison for ordinal data
comparison_poly <- data.frame(
  Model = c("GRM", "PCM", "GPCM"),
  LogLik = c(fit_grm$logLik, fit_pcm$logLik, fit_gpcm$logLik),
  AIC = c(fit_grm$AIC, fit_pcm$AIC, fit_gpcm$AIC),
  BIC = c(fit_grm$BIC, fit_pcm$BIC, fit_gpcm$BIC)
)

print(comparison_poly)
# GRM typically fits better than PCM if discriminations vary
# PCM is more parsimonious (fewer parameters)
```

## Differential Item Functioning (DIF)

DIF analysis tests whether items function differently across groups (e.g., gender, ethnicity, language).

### Types of DIF

- **Uniform DIF**: Item difficulty differs between groups (parallel ICCs)
- **Non-uniform DIF**: Item discrimination differs between groups (non-parallel ICCs)

### Simulating Data with DIF

```{r dif-simulation}
set.seed(789)
n_persons <- 400
n_items <- 20

# Create two groups
group <- rep(c("Group A", "Group B"), each = n_persons/2)

# Generate abilities (equal distribution across groups)
theta <- rnorm(n_persons, 0, 1)

# Item parameters
difficulty <- rnorm(n_items, 0, 1)
discrimination <- runif(n_items, 0.8, 2.0)

# Add uniform DIF to items 5, 10, and 15 for Group B
dif_items <- c(5, 10, 15)
dif_amount <- c(0.8, 1.2, -0.6)

responses_dif <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    diff_effective <- difficulty[j]

    # Add DIF if applicable
    if (j %in% dif_items && group[i] == "Group B") {
      dif_idx <- which(dif_items == j)
      diff_effective <- diff_effective + dif_amount[dif_idx]
    }

    p <- plogis(discrimination[j] * (theta[i] - diff_effective))
    responses_dif[i, j] <- rbinom(1, 1, p)
  }
}
```

### Testing for DIF

```{r dif-test}
# Test all items for DIF
dif_result <- dif_test_with_data(
  responses_dif,
  group = group,
  model = "2PL",
  alpha = 0.05
)

print(dif_result)
```

### Interpreting DIF Results

```{r dif-interpret}
# View detailed results
summary(dif_result)

# Items flagged for DIF
cat("Flagged items:", dif_result$flagged_items, "\n")

# Effect sizes (ETS Delta scale)
dif_results_df <- dif_result$dif_results
flagged_df <- dif_results_df[dif_results_df$item %in% dif_result$flagged_items, ]
print(flagged_df)

# Effect size interpretation:
# |Delta| < 0.3: Negligible DIF
# 0.3 <= |Delta| < 0.5: Moderate DIF
# |Delta| >= 0.5: Large DIF
```

### Visualizing DIF

```{r dif-plot}
# Plot ICCs for an item showing DIF
dif_plot(dif_result, item = 5)
# Blue line: Group A, Red line: Group B
# Large separation indicates DIF
```

### DIF with Polytomous Items

```{r dif-polytomous}
# DIF analysis also works with polytomous models
dif_result_grm <- dif_test_with_data(
  responses_grm,
  group = sample(c("Male", "Female"), nrow(responses_grm), replace = TRUE),
  model = "GRM",
  alpha = 0.05
)

print(dif_result_grm)
```

## Explanatory IRT (EIRT)

Explanatory IRT models item parameters as functions of item characteristics (covariates).

### Motivation

Instead of estimating $b_j$ and $a_j$ separately for each item, model them as:

$$b_j = \gamma_0 + \gamma_1 W_{j1} + \gamma_2 W_{j2} + ... + \epsilon_{bj}$$
$$\log(a_j) = \delta_0 + \delta_1 V_{j1} + \delta_2 V_{j2} + ... + \epsilon_{aj}$$

where $W$ and $V$ are item-level covariates.

### Example: Reading Test Items

```{r eirt-setup}
set.seed(111)
n_persons <- 300
n_items <- 25

# Item characteristics
item_data <- data.frame(
  word_frequency = rnorm(n_items, 0, 1),      # Higher = more common words
  passage_length = rpois(n_items, 100),       # Length in words
  item_type = factor(sample(c("Literal", "Inferential", "Critical"),
                             n_items, replace = TRUE))
)

# True covariate effects
gamma_0 <- 0.0          # Intercept
gamma_freq <- -0.6      # Higher word freq → easier (lower difficulty)
gamma_length <- 0.003   # Longer passages → harder

delta_0 <- 0.3          # Log-discrimination intercept
delta_type_inf <- 0.4   # Inferential items discriminate better
delta_type_crit <- 0.6  # Critical thinking items discriminate even better

# Generate true item parameters from covariates
theta <- rnorm(n_persons, 0, 1)

difficulty <- gamma_0 +
              gamma_freq * item_data$word_frequency +
              gamma_length * item_data$passage_length +
              rnorm(n_items, 0, 0.3)  # residual variation

discrimination <- exp(delta_0 +
                     delta_type_inf * (item_data$item_type == "Inferential") +
                     delta_type_crit * (item_data$item_type == "Critical") +
                     rnorm(n_items, 0, 0.2))

# Generate responses
responses_eirt <- matrix(NA, n_persons, n_items)
for (i in 1:n_persons) {
  for (j in 1:n_items) {
    p <- plogis(discrimination[j] * (theta[i] - difficulty[j]))
    responses_eirt[i, j] <- rbinom(1, 1, p)
  }
}
```

### Fitting EIRT Models

```{r fit-eirt}
# Fit EIRT with item covariates
fit_eirt <- fit_eirt(
  responses_eirt,
  item_data = item_data,
  difficulty_formula = ~ word_frequency + passage_length,
  discrimination_formula = ~ item_type,
  model = "2PL"
)

print(fit_eirt)
```

### Interpreting EIRT Results

```{r eirt-results}
# Difficulty regression coefficients
gamma_hat <- fit_eirt$regression_coefficients$difficulty
print("Difficulty regression:")
print(gamma_hat)

# Interpretation:
# - Negative word_frequency: items with common words are easier
# - Positive passage_length: longer passages are harder

# Discrimination regression coefficients
delta_hat <- fit_eirt$regression_coefficients$discrimination
print("Discrimination regression:")
print(delta_hat)

# Interpretation:
# - Positive coefficients for Inferential/Critical:
#   these item types discriminate better than Literal items

# Residual standard deviations
print("Residual SDs:")
print(fit_eirt$residual_sd)
# How much variation remains after accounting for covariates
```

### Comparing EIRT to Standard IRT

```{r eirt-comparison}
# Standard IRT (no covariates)
fit_standard <- fit_irt(responses_eirt, model = "2PL")

# EIRT (with covariates)
# fit_eirt already fitted above

# Model comparison
cat("Standard IRT:\n")
cat("  Log-likelihood:", fit_standard$logLik, "\n")
cat("  AIC:", fit_standard$AIC, "\n")

cat("\nExplanatory IRT:\n")
cat("  Log-likelihood:", fit_eirt$logLik, "\n")
cat("  AIC:", fit_eirt$AIC, "\n")

# EIRT may have better AIC if covariates explain substantial variation
# EIRT provides interpretable regression coefficients
```

### EIRT with Polytomous Items

```{r eirt-polytomous}
# Create item characteristics for Likert-scale items
item_data_likert <- data.frame(
  complexity = rnorm(n_items, 0, 1),
  reverse_coded = factor(sample(c("Yes", "No"), n_items, replace = TRUE))
)

# Fit EIRT with GRM
fit_eirt_grm <- fit_eirt(
  responses_grm,
  item_data = item_data_likert,
  difficulty_formula = ~ complexity + reverse_coded,
  discrimination_formula = ~ complexity,
  model = "GRM"
)

print(fit_eirt_grm)
summary(fit_eirt_grm)
```

### Practical Applications of EIRT

#### Test Development

```{r test-development, eval=FALSE}
# Use EIRT to:
# 1. Identify item characteristics that predict difficulty
# 2. Design new items with targeted difficulty
# 3. Balance tests by controlling item characteristics

# Example: Create a new item
new_item <- data.frame(
  word_frequency = 0.5,  # Moderately common words
  passage_length = 90,   # Short passage
  item_type = "Literal"
)

# Predict difficulty and discrimination
W_new <- model.matrix(~ word_frequency + passage_length, data = new_item)
gamma <- fit_eirt$regression_coefficients$difficulty
predicted_difficulty <- sum(W_new * gamma)

cat("Predicted difficulty for new item:", predicted_difficulty, "\n")
```

#### Item Revision

```{r item-revision, eval=FALSE}
# Identify items with large residuals (not well-explained by covariates)
item_params <- fit_eirt$item_parameters
predicted_diffs <- model.matrix(~ word_frequency + passage_length,
                                data = item_data) %*%
                   fit_eirt$regression_coefficients$difficulty

residuals_diff <- item_params$difficulty - predicted_diffs

# Items with large positive residuals are unexpectedly hard
problematic_items <- which(abs(residuals_diff) > 1)
cat("Items to review:", problematic_items, "\n")
```

## Advanced Topics

### Multidimensional IRT

```{r mirt, eval=FALSE}
# Future feature: Multiple latent dimensions
# Useful for tests measuring multiple constructs
# fit_mirt <- fit_irt(responses, model = "2PL", dimensions = 2)
```

### Computerized Adaptive Testing (CAT)

```{r cat, eval=FALSE}
# Future feature: Item selection algorithms
# Select next item based on current ability estimate
# next_item <- select_item_cat(fit_irt, current_ability, method = "MFI")
```

## References

- Rasch, G. (1960). Probabilistic Models for Some Intelligence and Attainment Tests.
- Lord, F. M. (1980). Applications of Item Response Theory to Practical Testing Problems.
- Embretson, S. E., & Reise, S. P. (2000). Item Response Theory for Psychologists.
- Baker, F. B., & Kim, S. H. (2004). Item Response Theory: Parameter Estimation Techniques.

## See Also

- `vignette("latent-class")` - Latent class models
- `vignette("factor-analysis")` - Factor analysis with GLLAMMR
- `vignette("mixed-response")` - Mixed response models
