---
title: "Latent Class Analysis with GLLAMMR"
author: "GLLAMMR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent Class Analysis with GLLAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction to Latent Class Analysis

Note: the code in this vignette is illustrative and is not evaluated when the vignette is built.

Latent Class Analysis (LCA) is a statistical method for identifying unobserved subgroups (latent classes) within a population based on responses to categorical variables.

### Key Concepts

- **Latent classes**: Unobserved subgroups in the population
- **Manifest variables**: Observed categorical indicators
- **Class membership probabilities**: Proportion of population in each class
- **Item response probabilities**: Probability of each response given class membership
- **Local independence**: Within classes, responses are independent

## Basic LCA Model

The LCA model for binary items:

$$P(Y_i = y) = \sum_{k=1}^K \pi_k \prod_{j=1}^J \rho_{jk}^{y_{ij}} (1-\rho_{jk})^{1-y_{ij}}$$

where:
- $K$ is the number of latent classes
- $\pi_k$ is the probability of class $k$
- $\rho_{jk}$ is the probability of endorsing item $j$ in class $k$

## Simulating LCA Data

```{r simulate-lca}
library(GLLAMMR)

set.seed(123)
n <- 500
n_items <- 6

# Two latent classes with distinct response patterns
# Class 1: "High endorsers" - frequently say yes
class1_probs <- c(0.85, 0.75, 0.90, 0.80, 0.85, 0.70)

# Class 2: "Low endorsers" - infrequently say yes
class2_probs <- c(0.15, 0.25, 0.10, 0.20, 0.15, 0.30)

# 60% in class 1, 40% in class 2
true_class <- sample(1:2, n, replace = TRUE, prob = c(0.6, 0.4))

# Generate item responses
data <- matrix(NA, n, n_items)
for (i in 1:n) {
  probs <- if (true_class[i] == 1) class1_probs else class2_probs
  data[i, ] <- rbinom(n_items, 1, probs)
}

colnames(data) <- paste0("Item", 1:n_items)
head(data)
```

## Fitting a 2-Class Model

```{r fit-2class}
fit_2class <- fit_lca(data, nclass = 2)

print(fit_2class)
```

The output shows:
- Class probabilities (proportion in each class)
- Item response probabilities for each class
- Model fit statistics (AIC, BIC, log-likelihood)

## Examining Results

### Class Sizes

```{r class-sizes}
# Modal class assignment
table(fit_2class$modal_class)

# Compare to true classes
if (exists("true_class")) {
  # Confusion matrix (may need to relabel)
  table(True = true_class, Estimated = fit_2class$modal_class)

  # Classification accuracy (best of two label permutations)
  acc1 <- mean(fit_2class$modal_class == true_class)
  acc2 <- mean(fit_2class$modal_class == (3 - true_class))
  cat("Classification accuracy:", max(acc1, acc2), "\n")
}
```

### Item Response Probabilities

```{r item-probs}
# Item probabilities by class
print(round(fit_2class$item_probs, 3))

# Visualize item profiles
barplot(t(fit_2class$item_probs),
        beside = TRUE,
        legend.text = paste("Class", 1:2),
        xlab = "Item",
        ylab = "P(Endorse)",
        main = "Item Response Profiles by Class",
        col = c("lightblue", "lightcoral"))
```

### Posterior Probabilities

```{r posterior}
# First few individuals' posterior class probabilities
head(fit_2class$posterior)

# How certain are we about class membership?
# High max posterior = high certainty
max_posterior <- apply(fit_2class$posterior, 1, max)

hist(max_posterior,
     main = "Distribution of Maximum Posterior Probabilities",
     xlab = "Max Posterior Probability",
     col = "lightgreen",
     breaks = 20)

# Individuals with uncertain classification
uncertain <- which(max_posterior < 0.7)
cat("Number of uncertain classifications:", length(uncertain), "\n")
```

## Determining the Number of Classes

A key question in LCA is how many classes exist. Use information criteria and interpretability.

```{r model-selection}
# Fit models with different numbers of classes
fit_1class <- fit_lca(data, nclass = 1)  # Null model
fit_2class <- fit_lca(data, nclass = 2)
fit_3class <- fit_lca(data, nclass = 3)
fit_4class <- fit_lca(data, nclass = 4)

# Compare fit statistics
model_comparison <- data.frame(
  nclass = 1:4,
  LogLik = c(fit_1class$logLik, fit_2class$logLik,
             fit_3class$logLik, fit_4class$logLik),
  AIC = c(fit_1class$AIC, fit_2class$AIC,
          fit_3class$AIC, fit_4class$AIC),
  BIC = c(fit_1class$BIC, fit_2class$BIC,
          fit_3class$BIC, fit_4class$BIC)
)

print(model_comparison)

# Plot BIC by number of classes
plot(model_comparison$nclass, model_comparison$BIC,
     type = "b", lwd = 2, pch = 19,
     xlab = "Number of Classes",
     ylab = "BIC",
     main = "Model Selection via BIC")

# Lowest BIC is preferred
best_model <- which.min(model_comparison$BIC)
cat("Best model (by BIC):", best_model, "classes\n")
```

## Interpreting Classes

### Naming Classes

Based on item profiles, assign meaningful names:

```{r interpret}
# Example: Assuming fit_2class is the best model
class_profiles <- fit_2class$item_probs

# Class 1: High on all items -> "High Symptom" group
# Class 2: Low on all items -> "Low Symptom" group

class_names <- c("High Symptom", "Low Symptom")

# Or for more nuanced patterns:
# Class 1: High on items 1-3, low on 4-6 -> "Anxious" subtype
# Class 2: Low on items 1-3, high on 4-6 -> "Depressed" subtype
```

### Profiling Classes

```{r profiling, eval=FALSE}
# If you have additional covariates (not used in LCA)
# you can profile the classes

# Example: Are classes related to age or gender?
# (Assuming we have these variables)

# age_by_class <- tapply(age, fit_2class$modal_class, mean)
# gender_by_class <- table(gender, fit_2class$modal_class)
```

## Practical Applications

### Clinical Psychology: Depression Subtypes

```{r depression-example, eval=FALSE}
# Simulate depression symptom data
set.seed(456)
n_patients <- 600
symptoms <- c("Sad", "Tired", "Insomnia", "Appetite", "Guilt",
              "Concentration", "Psychomotor", "Suicidal", "Anhedonia")
n_symptoms <- length(symptoms)

# Three classes:
# 1. Severe (high on all)
# 2. Moderate (medium on most)
# 3. Mild (low on most)

severe_probs <- rep(0.85, n_symptoms)
moderate_probs <- rep(0.50, n_symptoms)
mild_probs <- rep(0.15, n_symptoms)

true_severity <- sample(1:3, n_patients, replace = TRUE,
                        prob = c(0.2, 0.5, 0.3))

depression_data <- matrix(NA, n_patients, n_symptoms)
for (i in 1:n_patients) {
  probs <- switch(true_severity[i],
                  severe_probs,
                  moderate_probs,
                  mild_probs)
  depression_data[i, ] <- rbinom(n_symptoms, 1, probs)
}

colnames(depression_data) <- symptoms

# Fit 3-class model
fit_depression <- fit_lca(depression_data, nclass = 3)

print(fit_depression)

# Examine class profiles
print(round(fit_depression$item_probs, 2))
```

### Education: Learning Strategies

```{r learning-strategies, eval=FALSE}
# Simulate data on study strategies
set.seed(789)
n_students <- 400
strategies <- c("Highlighting", "Summarizing", "Practice_Testing",
                "Distributed", "Interleaved", "Elaboration")
n_strategies <- length(strategies)

# Two classes:
# 1. "Deep learners" - use effective strategies
# 2. "Surface learners" - use ineffective strategies

deep_probs <- c(0.3, 0.8, 0.9, 0.85, 0.7, 0.8)  # High on good strategies
surface_probs <- c(0.9, 0.3, 0.2, 0.15, 0.2, 0.3)  # High on highlighting only

true_type <- sample(1:2, n_students, replace = TRUE, prob = c(0.4, 0.6))

strategy_data <- matrix(NA, n_students, n_strategies)
for (i in 1:n_students) {
  probs <- if (true_type[i] == 1) deep_probs else surface_probs
  strategy_data[i, ] <- rbinom(n_strategies, 1, probs)
}

colnames(strategy_data) <- strategies

# Fit model
fit_strategies <- fit_lca(strategy_data, nclass = 2)

print(fit_strategies)

# Interpret classes
print(round(fit_strategies$item_probs, 2))

# Deep learners should show high use of:
# - Practice testing
# - Distributed practice
# - Interleaving
# - Elaboration
```

### Marketing: Customer Segmentation

```{r marketing, eval=FALSE}
# Simulate customer purchase patterns
set.seed(111)
n_customers <- 800
products <- c("Premium", "Budget", "Frequent", "Bulk",
              "Online", "Loyalty", "Reviews")
n_products <- length(products)

# Three customer segments:
# 1. "Premium shoppers"
# 2. "Budget hunters"
# 3. "Occasional buyers"

premium_probs <- c(0.9, 0.1, 0.8, 0.3, 0.9, 0.8, 0.7)
budget_probs <- c(0.1, 0.9, 0.6, 0.8, 0.5, 0.4, 0.8)
occasional_probs <- c(0.3, 0.5, 0.2, 0.1, 0.4, 0.2, 0.3)

true_segment <- sample(1:3, n_customers, replace = TRUE,
                       prob = c(0.25, 0.45, 0.30))

customer_data <- matrix(NA, n_customers, n_products)
for (i in 1:n_customers) {
  probs <- switch(true_segment[i],
                  premium_probs,
                  budget_probs,
                  occasional_probs)
  customer_data[i, ] <- rbinom(n_products, 1, probs)
}

colnames(customer_data) <- products

# Fit model
fit_customers <- fit_lca(customer_data, nclass = 3)

print(fit_customers)

# Target marketing strategies based on segments
print(round(fit_customers$item_probs, 2))
```

## Model Diagnostics

### Local Independence

Check if items are independent within classes (LCA assumption):

```{r local-indep, eval=FALSE}
# Compare observed and expected cell frequencies
# High residuals suggest local dependence

# This would require additional functionality:
# check_local_independence(fit_2class)
```

### Classification Diagnostics

```{r classification-diag}
# Entropy: measure of classification uncertainty
# Ranges from 0 (complete uncertainty) to 1 (perfect certainty)

compute_entropy <- function(posterior) {
  # Avoid log(0)
  posterior[posterior == 0] <- 1e-10

  # Entropy formula
  entropy_i <- -rowSums(posterior * log(posterior))
  max_entropy <- log(ncol(posterior))

  # Relative entropy (0 to 1)
  rel_entropy <- 1 - mean(entropy_i) / max_entropy

  return(rel_entropy)
}

entropy <- compute_entropy(fit_2class$posterior)
cat("Entropy:", round(entropy, 3), "\n")

# Entropy > 0.8 is considered good separation
```

## Advanced Topics

### Latent Class Regression

```{r lcr, eval=FALSE}
# Future feature: Include covariates predicting class membership
# fit_lcr <- fit_lca(data, nclass = 2, formula = ~ age + gender)
```

### Growth Mixture Models

```{r gmm, eval=FALSE}
# Future feature: Combine LCA with growth curves
# fit_gmm <- fit_growth_mixture(y ~ time | id, nclass = 3)
```

### Latent Class IRT

```{r lc-irt, eval=FALSE}
# Future feature: IRT models with discrete ability distribution
# fit_lc_irt <- fit_irt(responses, nclass = 3)
```

## Tips for Applied Researchers

1. **Start simple**: Begin with 2 classes, then increase
2. **Use multiple criteria**: Don't rely on fit statistics alone
3. **Check interpretability**: Classes should make substantive sense
4. **Validate**: Use holdout samples or cross-validation
5. **Report uncertainty**: Include entropy and posterior probabilities
6. **Consider alternatives**: Mixture models, hierarchical clustering

## References

- Lazarsfeld, P. F., & Henry, N. W. (1968). Latent Structure Analysis.
- McCutcheon, A. L. (1987). Latent Class Analysis.
- Hagenaars, J. A., & McCutcheon, A. L. (2002). Applied Latent Class Analysis.
- Collins, L. M., & Lanza, S. T. (2010). Latent Class and Latent Transition Analysis.
- Nylund-Gibson, K., & Choi, A. Y. (2018). Ten frequently asked questions about latent class analysis.

## See Also

- `vignette("irt-models")` - Item response theory
- `vignette("factor-analysis")` - Continuous latent variables
- `vignette("mixed-response")` - Multiple outcome types
