---
title: "Latent Class Analysis with gllammr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent Class Analysis with gllammr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

Latent class analysis (LCA) explains associations among categorical
indicators through membership in a small number of unobserved classes.
In the GLLAMM framework it is the discrete-latent-variable member of the
family: replace the normal latent trait with K mass points and the model
is an LCA.

`fit_lca()` (equivalently `gllamm(Y, family = lca(...))`) handles
binary, polytomous, and continuous (gaussian) indicators in any mix,
missing responses under MAR, case weights, and — beyond standard LCA —
**order-restricted** and **partially ordered** class structures.
Estimation is accelerated closed-form EM, validated against poLCA
(identical log-likelihoods on the classic carcinoma data and 20,000-case
simulations, at 2–3x poLCA's speed).

## A basic two-class model

```{r simulate}
library(gllammr)
set.seed(2026)
n <- 800
true_class <- rbinom(n, 1, 0.4) + 1
p_yes <- rbind(c(.80, .70, .85, .75, .65),   # class 1: high endorsers
               c(.20, .25, .15, .30, .25))   # class 2: low endorsers
Y <- sapply(1:5, function(j) rbinom(n, 1, p_yes[true_class, j]))
colnames(Y) <- paste0("sym", 1:5)

fit2 <- fit_lca(Y, nclass = 2)
print(fit2)
```

Class-membership posteriors and modal assignments support downstream
profiling; entropy summarizes how cleanly the classes separate:

```{r posterior}
post <- fit2$posterior
entropy <- 1 - sum(-post * log(pmax(post, 1e-12))) /
  (nrow(post) * log(fit2$nclass))
round(c(entropy = entropy,
        accuracy = max(mean(fit2$modal_class == true_class),
                       mean(fit2$modal_class == 3 - true_class))), 3)
```

## How many classes?

Fit a sequence and compare information criteria (BIC is the usual
default for LCA):

```{r enumeration}
fits <- lapply(2:4, function(k) fit_lca(Y, nclass = k,
                                        control = list(n_starts = 5)))
data.frame(classes = 2:4,
           logLik = sapply(fits, `[[`, "logLik"),
           BIC = sapply(fits, `[[`, "BIC"))
```

BIC correctly favours the two-class structure that generated the data.

## Ordered latent classes

When classes represent *grades* of a single phenomenon — severity,
proficiency, engagement — an unrestricted LCA leaves their ordering to
chance and label switching. `ordering = "increasing"` fits Croon's
ordered latent class model: every item probability (and every gaussian
indicator mean) is constrained nondecreasing from class 1 to class K.
The constrained M-step is a weighted isotonic regression, so estimation
stays closed-form.

```{r ordered}
set.seed(7)
cls3 <- sample(1:3, 900, TRUE, prob = c(.4, .35, .25))
pmat <- rbind(c(.10, .50, .90), c(.20, .40, .80), c(.15, .55, .85),
              c(.30, .50, .70), c(.05, .45, .90), c(.25, .60, .95))
Y3 <- sapply(1:6, function(j) rbinom(900, 1, pmat[j, cls3]))

ord <- fit_lca(Y3, nclass = 3, ordering = "increasing")
round(ord$item_probs, 3)
```

Every row increases left to right by construction: class labels are
pinned, and the model is the formal statement of "these classes are
ordered". Comparing its log-likelihood with the unrestricted fit tests
whether the ordering story is tenable (note the likelihood-ratio null is
chi-bar-square, so use the comparison descriptively):

```{r ordered-compare}
unr <- fit_lca(Y3, nclass = 3, control = list(n_starts = 5))
c(unrestricted = unr$logLik, ordered = ord$logLik)
```

A negligible gap (as here, where the truth is ordered) supports the
restriction.

## Partially ordered classes

Sometimes classes are ordered but not *linearly* — two intermediate
profiles may each dominate a low class and be dominated by a high class
while being incomparable with each other (mastered skill A vs mastered
skill B). Pass the comparable pairs:

```{r poset}
set.seed(61)
cls4 <- sample(1:4, 1200, TRUE)
# items 1-3 respond to profile 2; items 4-6 to profile 3
pmat4 <- rbind(
  c(.10, .80, .15, .90), c(.15, .85, .20, .90), c(.10, .75, .10, .85),
  c(.10, .15, .80, .90), c(.15, .20, .85, .90), c(.10, .10, .75, .85))
Y4 <- sapply(1:6, function(j) rbinom(1200, 1, pmat4[j, cls4]))

diamond <- fit_lca(Y4, nclass = 4,
                   ordering = list(c(1, 2), c(1, 3), c(2, 4), c(3, 4)))
round(diamond$item_probs, 3)

chain <- fit_lca(Y4, nclass = 4, ordering = "increasing",
                 control = list(n_starts = 5))
c(diamond = diamond$logLik, total_order = chain$logLik)
```

The diamond fits essentially as well as an unrestricted model, while
*no* total order can accommodate the crossing intermediate profiles —
which is precisely the diagnostic logic: if `"increasing"` costs a lot
but your hypothesized lattice does not, the classes are ordered but not
linearly. Fully Q-matrix-structured class spaces are the next step up:
see `vignette("cognitive-diagnosis")`.

## Mixed indicator types and missing data

Indicators are auto-detected: 0/1 columns are binary, integer 1..K
columns categorical, anything else gaussian. Missing responses are
handled by the EM under MAR — each person contributes whatever they
answered:

```{r mixed}
Ym <- cbind(Y3[, 1:4],
            score = rnorm(900, mean = c(-1, 0, 1)[cls3], sd = .8))
Ym[sample(length(Ym), 300)] <- NA
mix <- fit_lca(Ym, nclass = 3, ordering = "increasing")
round(mix$gaussian_params$means, 2)
```

The ordering restriction applies to the gaussian means as well — the
class with the higher symptom probabilities also has the higher mean
score, by construction.

## Practical guidance

- Use several random starts (`control = list(n_starts = ...)`) for
  unrestricted models; ordering reduces the label-switching part of the
  multimodality problem but not the rest.
- Profile classes against external variables using `posterior` weights
  rather than modal assignment when entropy is modest.
- Inequality-restricted models report AIC/BIC with the nominal parameter
  count; treat comparisons against unrestricted models as descriptive.

## References

- Croon, M. (1990). Latent class analysis with ordered latent classes.
  *British Journal of Mathematical and Statistical Psychology*, 43,
  171–192.
- Linzer, D. A., & Lewis, J. B. (2011). poLCA: an R package for
  polytomous variable latent class analysis. *Journal of Statistical
  Software*, 42(10).
- Skrondal, A., & Rabe-Hesketh, S. (2004). *Generalized Latent Variable
  Modeling*. Chapman & Hall/CRC.
