---
title: "Marginal vs Conditional Predictions in GLLAMMR"
author: "GLLAMMR Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal vs Conditional Predictions in GLLAMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette explains the difference between **conditional** and **marginal** predictions in generalized linear mixed models (GLMMs) and other hierarchical models, and demonstrates how to compute both types of predictions using GLLAMMR.

## What are Marginal Predictions?

In mixed models, we distinguish between two types of predictions:

### Conditional Predictions

Conditional predictions are specific to a particular level of the random effects. When we set the random effects to zero (u = 0), we get predictions at the "average" cluster level:

```
E[Y | X, u = 0] = g^(-1)(X'β)
```

where g^(-1) is the inverse link function.

### Marginal Predictions

Marginal predictions are averaged over the distribution of random effects, giving us **population-level** predictions:

```
E[Y | X] = ∫ g^(-1)(X'β + Z'u) f(u) du
```

where u ~ N(0, Σ_u).

### Why They Differ

For **nonlinear link functions** (logit, probit, log), marginal predictions differ from conditional predictions due to Jensen's inequality:

```
E[g^(-1)(X'β + Z'u)] ≠ g^(-1)(X'β)
```

For the **identity link** (Gaussian models), they are equal:

```
E[X'β + Z'u] = X'β
```

## Computing Marginal Predictions

GLLAMMR uses Monte Carlo integration to compute marginal predictions:

1. Draw S samples from the random effects distribution: u_s ~ N(0, Σ_u)
2. For each sample, compute: ŷ_s = g^(-1)(X'β + Z'u_s)
3. Average: E[Y|X] ≈ (1/S) Σ_s ŷ_s

Default: S = 1000 samples

## Example 1: Binomial GLMM

Let's start with a simple example where the difference between conditional and marginal predictions is clear.

```{r binomial_example, eval=FALSE}
library(GLLAMMR)
set.seed(123)

# Simulate data: treatment effect with clinic random effects
n_clinics <- 20
n_per_clinic <- 25
n <- n_clinics * n_per_clinic

data <- data.frame(
  treatment = rep(0:1, each = n/2),
  clinic = factor(rep(1:n_clinics, each = n_per_clinic)),
  success = rbinom(n, 1, prob = 0.6)
)

# Fit binomial GLMM
fit <- gllamm(success ~ treatment + (1 | clinic),
              data = data,
              family = binomial(link = "logit"))

# Conditional predictions (at u = 0)
pred_cond <- predict(fit, type = "response")

# Marginal predictions (averaged over clinics)
pred_marg <- predict(fit, type = "marginal", n_sim = 1000)

# Compare
summary(pred_cond - pred_marg)
```

**Key insight:** Marginal predictions are typically lower than conditional predictions for logit/probit links when random effects have substantial variance. This is because the inverse logit is a concave function.

```{r plot_comparison, eval=FALSE}
# Visualize the difference
plot(pred_cond, pred_marg,
     xlab = "Conditional Prediction",
     ylab = "Marginal Prediction",
     main = "Conditional vs Marginal Predictions",
     pch = 19, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red", lwd = 2)
```

## Example 2: Ordinal Models

For ordinal outcomes, marginal predictions give us population-level category probabilities.

```{r ordinal_example, eval=FALSE}
# Simulate ordinal data
data_ord <- data.frame(
  rating = sample(1:5, n, replace = TRUE),
  temp = rnorm(n),
  judge = factor(rep(1:n_clinics, each = n_per_clinic))
)

# Fit proportional odds model
fit_ord <- gllamm(rating ~ temp + (1 | judge),
                  data = data_ord,
                  family = ordinal(link = "logit"))

# Marginal category probabilities
pred_marg_ord <- predict(fit_ord, type = "marginal", n_sim = 1000)

# Average marginal probabilities
colMeans(pred_marg_ord)
# This gives population-level probability for each rating category
```

**Interpretation:** These are the expected proportions of each rating category in the population, accounting for between-judge variability.

## Example 3: IRT Models

In IRT, marginal predictions give us the proportion of examinees who would answer each item correctly.

```{r irt_example, eval=FALSE}
# Simulate item response data
n_persons <- 200
n_items <- 10

responses <- matrix(rbinom(n_persons * n_items, 1, 0.6),
                   n_persons, n_items)

# Fit 2PL model
fit_irt <- fit_irt(responses, model = "2PL")

# Marginal item response probabilities
pred_marg_irt <- predict(fit_irt, type = "marginal", n_sim = 1000)

# Compare to empirical proportions
emp_props <- colMeans(responses, na.rm = TRUE)

# Plot
plot(emp_props, pred_marg_irt,
     xlab = "Empirical Proportion Correct",
     ylab = "Predicted Marginal Probability",
     main = "IRT: Empirical vs Predicted",
     pch = 19, cex = 1.5, col = "blue")
abline(0, 1, col = "red", lwd = 2)
```

**Key insight:** Marginal IRT predictions tell us what proportion of the population would answer each item correctly, which is useful for:
- Item difficulty interpretation
- Test score distributions
- Population-level assessment

## Example 4: Explanatory IRT (EIRT)

EIRT allows predictions for **new items** based on item characteristics.

```{r eirt_example, eval=FALSE}
# Item data with predictors
item_data <- data.frame(
  item_id = 1:n_items,
  item_length = rnorm(n_items),
  content_area = factor(sample(c("Algebra", "Geometry"), n_items, replace = TRUE))
)

# Fit EIRT model
fit_eirt <- fit_eirt(responses,
                     item_data = item_data,
                     difficulty_formula = ~ item_length + content_area,
                     model = "2PL")

# Predict for new items
new_items <- data.frame(
  item_length = c(-1, 0, 1),
  content_area = factor(c("Algebra", "Algebra", "Geometry"))
)

# Marginal predictions for new items
pred_new <- predict(fit_eirt,
                   newdata = new_items,
                   type = "marginal",
                   n_sim = 1000)

print(data.frame(
  item_length = new_items$item_length,
  content_area = new_items$content_area,
  pred_prob = round(pred_new, 3)
))
```

**Use case:** This is powerful for:
- Automatic item generation
- Item bank calibration
- Predicting item performance before pilot testing

## Example 5: Multinomial Models

For multinomial outcomes, marginal predictions give population-level category probabilities.

```{r multinomial_example, eval=FALSE}
# Simulate multinomial data (3 categories)
data_mult <- data.frame(
  choice = sample(0:2, n, replace = TRUE),
  price = rnorm(n),
  quality = rnorm(n),
  person = factor(rep(1:100, length.out = n))
)

# Fit multinomial logit model
fit_mult <- fit_multinomial(choice ~ price + quality + (1 | person),
                            data = data_mult)

# Marginal category probabilities
pred_marg_mult <- predict(fit_mult, type = "marginal", n_sim = 1000)

# Average across population
pop_shares <- colMeans(pred_marg_mult)
print(round(pop_shares, 3))
```

**Interpretation:** Population-level market shares or choice probabilities.

## Example 6: Survival Models

Marginal survival curves show population-average survival.

```{r survival_example, eval=FALSE}
# Simulate survival data
data_surv <- data.frame(
  time = rexp(n, rate = 0.1),
  status = rbinom(n, 1, 0.7),
  treatment = rep(0:1, each = n/2),
  hospital = factor(rep(1:20, each = n/20))
)

# Fit Weibull survival model
fit_surv <- fit_survival(time ~ treatment + (1 | hospital),
                        data = data_surv,
                        event = data_surv$status,
                        distribution = "Weibull")

# Marginal survival curves
times <- c(5, 10, 15, 20, 25)
surv_marg <- predict(fit_surv,
                    type = "marginal_survival",
                    times = times,
                    n_sim = 1000)

# Average survival curve
avg_surv <- colMeans(surv_marg)

# Plot
plot(times, avg_surv, type = "b",
     xlab = "Time", ylab = "Survival Probability",
     main = "Population-Average Survival Curve",
     ylim = c(0, 1), pch = 19, col = "blue", lwd = 2)
grid()
```

## When to Use Marginal vs Conditional Predictions

### Use Conditional Predictions When:

1. **Cluster-specific inference**: You want to know the expected outcome for a specific cluster (e.g., a particular school, hospital, or person)
2. **Comparing clusters**: You're interested in between-cluster differences
3. **Prediction for new observation in existing cluster**: You have a new observation from a known cluster

### Use Marginal Predictions When:

1. **Population-level inference**: You want to know the expected outcome averaged across all clusters
2. **Policy evaluation**: Estimating population-level treatment effects
3. **Standardization**: Comparing populations with different cluster compositions
4. **New cluster prediction**: Predicting for observations from a new, unseen cluster
5. **Reporting to non-technical audiences**: Marginal predictions often have simpler interpretation

## Technical Details

### Computational Considerations

- **Default n_sim = 1000**: Good balance of accuracy and speed
- **Increase for final estimates**: Use n_sim = 5000 for publication
- **Decrease for exploration**: Use n_sim = 100-200 during development

Approximate computation times (n_obs = 1000, single random effect):

| Model Type | n_sim = 100 | n_sim = 1000 | n_sim = 5000 |
|------------|-------------|--------------|--------------|
| GLMM       | < 1 sec     | ~2 sec       | ~10 sec      |
| Ordinal    | ~1 sec      | ~3 sec       | ~15 sec      |
| IRT        | ~2 sec      | ~5 sec       | ~25 sec      |

### Standard Errors

For GLMMs, you can request standard errors:

```{r se_example, eval=FALSE}
pred_with_se <- predict(fit, type = "marginal", n_sim = 1000, se.fit = TRUE)
# Returns list with $fit and $se.fit
```

Standard errors are computed from the Monte Carlo sample variance.

### Special Optimization

For Gaussian models with identity link, marginal = conditional (no MC needed):

```{r gaussian_optimization, eval=FALSE}
fit_gauss <- gllamm(y ~ x + (1 | group), data = data, family = gaussian())

# These are identical (and instant):
pred_fixed <- predict(fit_gauss, re.form = NA)
pred_marg <- predict(fit_gauss, type = "marginal")

max(abs(pred_fixed - pred_marg))  # ~0
```

## Supported Models

Marginal predictions are available for:

- ✅ **GLMM**: Gaussian, Binomial, Poisson (all links)
- ✅ **Ordinal**: Proportional odds, adjacent category, continuation ratio
- ✅ **IRT**: Rasch, 2PL, 3PL, GRM, PCM, GPCM, NRM
- ✅ **EIRT**: Rasch, 2PL with item covariates
- ✅ **Multinomial**: Multinomial logit
- ✅ **Survival**: Exponential, Weibull

## References

- Skrondal, A., & Rabe-Hesketh, S. (2009). Prediction in multilevel generalized linear models. *Journal of the Royal Statistical Society: Series A*, 172(3), 659-687.

- Hedeker, D., & Gibbons, R. D. (2006). *Longitudinal Data Analysis*. Wiley.

- Zeger, S. L., Liang, K. Y., & Albert, P. S. (1988). Models for longitudinal data: a generalized estimating equation approach. *Biometrics*, 1049-1060.

## Summary

- **Marginal predictions** provide population-level inference by averaging over random effects
- For **nonlinear links**, marginal ≠ conditional (due to Jensen's inequality)
- Use `type = "marginal"` with `predict()` for any GLLAMMR model
- Adjust `n_sim` to balance accuracy and computation time
- Choose marginal vs conditional based on your inference goals

For more information, see `?predict.gllamm` and related predict methods.
