---
title: "Structural Equation Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural Equation Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`fit_sem()` (equivalently `gllamm(d, family = sem(...))`) fits
structural equation models with continuous indicators: confirmatory
factor analysis, recursive path models among latent variables, MIMIC
models (latent variables regressed on observed covariates), and
full-information maximum likelihood for missing data — with the standard
fit-index battery and lavaan-matching results (the validation suite pins
log-likelihoods, estimates, standard errors, and CFI/TLI/RMSEA/SRMR
against lavaan).

Estimation is Wishart maximum likelihood on the sample covariance matrix
for complete data — cost independent of sample size — and a casewise
(missing-pattern) likelihood under FIML.

## Confirmatory factor analysis

Two correlated factors, three indicators each. The first indicator of
each factor is the marker (loading fixed at 1); exogenous factors covary
freely, as in lavaan.

```{r cfa}
library(GLLAMMR)
set.seed(2026)
n <- 500
f1 <- rnorm(n)
f2 <- 0.6 * f1 + sqrt(1 - 0.36) * rnorm(n)   # factor correlation 0.6
d <- data.frame(
  x1 = f1 + rnorm(n, 0, .6),  x2 = 0.8 * f1 + rnorm(n, 0, .6),
  x3 = 1.2 * f1 + rnorm(n, 0, .6),
  y1 = f2 + rnorm(n, 0, .5),  y2 = 0.9 * f2 + rnorm(n, 0, .5),
  y3 = 1.1 * f2 + rnorm(n, 0, .5))

cfa <- fit_sem(
  measurement = list(verbal = ~ x1 + x2 + x3,
                     quant  = ~ y1 + y2 + y3),
  data = d)
print(cfa)
```

The header reports the global fit: a non-significant chi-square, CFI/TLI
near 1, RMSEA near 0 (with its 90% interval), and a small SRMR all
indicate the two-factor structure reproduces the observed covariances.
`summary()` adds the full parameter table — estimates, standard errors,
z statistics, and the standardized (std.all) solution, where loadings
become indicator-factor correlations and `verbal~~quant` the factor
correlation:

```{r cfa-summary}
summary(cfa)
```

## Structural and MIMIC models

Regressions among latent variables, and of latent variables on observed
covariates, go in `structural`. Mixing the two is the MIMIC model:

```{r mimic}
d$ses <- rnorm(n)
f2b <- 0.5 * f1 + 0.3 * d$ses + rnorm(n, 0, .75)   # quant depends on ses
d$y1 <- f2b + rnorm(n, 0, .5)
d$y2 <- 0.9 * f2b + rnorm(n, 0, .5)
d$y3 <- 1.1 * f2b + rnorm(n, 0, .5)

mimic <- fit_sem(
  measurement = list(verbal = ~ x1 + x2 + x3,
                     quant  = ~ y1 + y2 + y3),
  structural = list(quant ~ verbal + ses),
  data = d)
mimic$param_table[grepl("~", mimic$param_table$label) &
                    !grepl("~~|=~", mimic$param_table$label), ]
```

Observed covariates are handled in the joint-normal formulation, so the
results are likelihood-equivalent to lavaan with `fixed.x = FALSE`.

## Missing data: FIML

The default is listwise deletion (with a warning). With
`missing = "fiml"` every observed value contributes through the
missing-pattern likelihood — the assumption is missing at random:

```{r fiml}
d_na <- d
set.seed(1)
d_na$x1[sample(n, 60)] <- NA
d_na$y2[sample(n, 60)] <- NA

fiml <- fit_sem(
  measurement = list(verbal = ~ x1 + x2 + x3,
                     quant  = ~ y1 + y2 + y3),
  structural = list(quant ~ verbal),
  data = d_na, missing = "fiml")
c(n_used = fiml$n_obs, converged = fiml$convergence$converged)
fiml$fit_measures[c("chisq", "df", "cfi", "rmsea")]
```

All 500 rows are retained, factor scores are computed casewise from each
person's observed variables, and the fit indices use an EM-estimated
saturated model. (One convention note: under FIML our SRMR uses the
EM-saturated covariance, which can differ from lavaan's in the third
decimal; everything else matches exactly.)

## Reading the fit measures

```{r fitm}
round(cfa$fit_measures, 4)
```

- `chisq`/`df`/`pvalue`: exact-fit test against the saturated model.
- `cfi`, `tli`: incremental fit vs the independence baseline (rules of
  thumb: > 0.95 good).
- `rmsea` with 90% CI: misfit per degree of freedom (< 0.06 good).
- `srmr`: average standardized covariance residual (< 0.08 good).

Inequality-free SEMs here are standard likelihood territory, so AIC/BIC
comparisons across non-nested structures are also reported.

## Notes on scope

Indicators are continuous; binary/ordinal measurement models belong to
the IRT side of the package (`vignette("irt-models")`), and joint models
mixing indicator types share latent variables through
`mixed_response()`. The structural model must be recursive (no feedback
loops). The legacy full-data Laplace path (`method = "laplace"`) remains
for latent-only models but treats exogenous factors as uncorrelated and
is no longer the default.

## References

- Bollen, K. A. (1989). *Structural Equations with Latent Variables*.
  Wiley.
- Skrondal, A., & Rabe-Hesketh, S. (2004). *Generalized Latent Variable
  Modeling*. Chapman & Hall/CRC.
- Rosseel, Y. (2012). lavaan: an R package for structural equation
  modeling. *Journal of Statistical Software*, 48(2).
